---
title: "Developmental Scaling"
---

# Preamble

## Install Libraries

```{r}
#install.packages("remotes")
#remotes::install_github("DevPsyLab/petersenlab")
```

## Load Libraries

```{r}
library("lavaan")
library("semTools")
library("semPlot")
library("lavaanPlot")
library("lavaangui")
library("sirt")
library("mirt")
library("plink")
library("rstan")
library("brms")
library("parallelly")
library("easystats")
library("kableExtra")
library("tidyverse")
```

## Import Data

```{r}
mydata_wide <- read.csv("./data/Bliese-Ployhart-2002-indicators-4.csv")
```

## Data Processing

```{r}
mydata_wide$sad1 <- mydata_wide$JOBSAT11
mydata_wide$sad2 <- mydata_wide$JOBSAT21
mydata_wide$sad3 <- mydata_wide$JOBSAT31

mydata_wide$down1 <- mydata_wide$JOBSAT12
mydata_wide$down2 <- mydata_wide$JOBSAT22
mydata_wide$down3 <- mydata_wide$JOBSAT32

mydata_wide$depressed1 <- mydata_wide$JOBSAT13
mydata_wide$depressed2 <- mydata_wide$JOBSAT23
mydata_wide$depressed3 <- mydata_wide$JOBSAT33
```

## Simulate Data

```{r}
#| code-fold: true

set.seed(52242)

N   <- 1200
rho <- 0.6                                 # person-level stability across time
Sigma <- rho ^ abs(outer(1:3, 1:3, "-"))   # AR(1) correlation: 1, rho, rho^2
mu <- c(-0.5, 0, 0.5)                      # increasing growth trend

theta_mat <- MASS::mvrnorm(N, mu = mu, Sigma = Sigma)
theta1_true <- theta_mat[, 1]
theta2_true <- theta_mat[, 2]
theta3_true <- theta_mat[, 3]

# simulate GRM responses
# theta : vector of person abilities
# a     : item discrimination (scalar)
# b     : vector of category thresholds, increasing (length = n categories - 1)
# returns integer responses coded 0 .. (n categories - 1)
sim_grm_item <- function(theta, a, b) {
  K <- length(b) + 1
  n <- length(theta)
  cumP <- matrix(NA_real_, n, K + 1)
  cumP[, 1] <- 1                                          # P(Y >= 0) = 1
  for (k in 1:(K - 1)) cumP[, k + 1] <- 1 / (1 + exp(-a * (theta - b[k])))
  cumP[, K + 1] <- 0                                       # P(Y >= K) = 0
  catP <- cumP[, 1:K] - cumP[, 2:(K + 1)]                  # category probabilities
  apply(catP, 1, function(p) sample.int(K, size = 1, prob = pmax(p, 1e-8))) - 1L
}

n_cat    <- 4     # every item has 4 ordered categories (3 thresholds)
n_unique <- 8     # unique ("age-specific") items at each timepoint
n_common <- 6     # common ("anchor") items linking each ADJACENT pair of timepoints

# true item parameters: a ~ U(0.8, 2), three increasing thresholds
make_params <- function(n_items) {
  a   <- round(runif(n_items, 0.8, 2.0), 2)
  b1  <- round(rnorm(n_items, -1.0, 0.5), 2)
  gap <- matrix(round(runif(n_items * 2, 0.5, 1.2), 2), n_items, 2)
  b   <- cbind(b1, b1 + gap[, 1], b1 + gap[, 1] + gap[, 2])
  list(a = a, b = b)
}

U1  <- make_params(n_unique)   # unique to T1
U2  <- make_params(n_unique)   # unique to T2
U3  <- make_params(n_unique)   # unique to T3
C12 <- make_params(n_common)   # anchor items shared by T1 & T2 (SAME true params both times)
C23 <- make_params(n_common)   # anchor items shared by T2 & T3

item_names_T1 <- c(paste0("U1_",  1:n_unique), paste0("C12_", 1:n_common))
item_names_T2 <- c(paste0("C12_", 1:n_common), paste0("U2_",  1:n_unique), paste0("C23_", 1:n_common))
item_names_T3 <- c(paste0("C23_", 1:n_common), paste0("U3_",  1:n_unique))

gen_data <- function(theta, names, params_list) {
  out <- matrix(NA_integer_, length(theta), length(names), dimnames = list(NULL, names))
  idx <- 1
  for (p in params_list) {
    for (j in seq_along(p$a)) {
      out[, idx] <- sim_grm_item(theta, p$a[j], p$b[j, ])
      idx <- idx + 1
    }
  }
  out
}

dat1 <- cbind(id = 1:N, as.data.frame(gen_data(theta1_true, item_names_T1, list(U1, C12))))
dat2 <- cbind(id = 1:N, as.data.frame(gen_data(theta2_true, item_names_T2, list(C12, U2, C23))))
dat3 <- cbind(id = 1:N, as.data.frame(gen_data(theta3_true, item_names_T3, list(C23, U3))))

# prepare data for concurrent calibration
all_items <- union(union(item_names_T1, item_names_T2), item_names_T3)

make_wide <- function(dat, all_items) {
  item_cols <- setdiff(colnames(dat), "id")
  out <- matrix(NA_integer_, nrow(dat), length(all_items), dimnames = list(NULL, all_items))
  out[, item_cols] <- as.matrix(dat[, item_cols])
  as.data.frame(out)
}

group <- factor(
  rep(c("T1", "T2", "T3"), each = N),
  levels = c("T1", "T2", "T3"))
id_stack <- c(dat1$id, dat2$id, dat3$id)   # same 1:N repeated 3x, in block order

dat_concurrent <- data.frame(
  id = id_stack,
  timepoint = group,
  rbind(make_wide(dat1, all_items), make_wide(dat2, all_items), make_wide(dat3, all_items))
)

# transform data to long format for Bayesian item response modeling
dat_long <- dat_concurrent |>
  pivot_longer(
    cols = all_of(all_items),
    names_to = "item",
    values_to = "response"
  ) |>
  filter(!is.na(response)) |>
  mutate(
    item = factor(item),
    response = factor(response, ordered = TRUE)
  ) |>
  arrange(id, timepoint, item)

dat_long$time <- NA
dat_long$time[which(dat_long$timepoint == "T1")] <- 1
dat_long$time[which(dat_long$timepoint == "T2")] <- 2
dat_long$time[which(dat_long$timepoint == "T3")] <- 3

dat_long$timeCentered <- dat_long$time - 2

dat_long$idFactor <- factor(dat_long$id)
dat_long$itemFactor <- factor(dat_long$item)
```

# Approaches

1. [Partial strong longitudinal factorial invariance](#sec-partialStrongInvariance) (paired with second-order growth functions—i.e., a second-order growth curve model)
1. [Alignment optimization](#sec-alignment) (paired with second-order growth functions)
1. [Moderated nonlinear factor analysis](#sec-mnlfa) (MNLFA) (paired with second-order growth functions)
1. [Item response theory (IRT) with anchor items](#sec-irt) (potentially with regularization/Bayesian penalty methods on the priors)
    - [separate calibration](#sec-irtSeparateCalibration)
    - [concurrent calibration](#sec-irtConcurrentCalibration)

# Partial Strong Longitudinal Factorial Invariance {#sec-partialStrongInvariance}

The example is adapted from the section on [longitudinal measurement invariance](#sec-longitudinalMI).

Two primary options for identifying a latent factor include:

1. standardizing the latent factor(s) at all timepoints (fixing their intercepts to zero and their variances to one)
1. standardizing the latent factor(s) at T1 (fixing their intercepts to zero and their variances to one) and constraining the factor loadings and intercepts of one item to be equal across time

(Option #2 may be preferable in alignment optimization.)

## Standardizing the Latent Factors at All Timepoints

## Standardizing the Latent Factors at T1 while Constraining Loadings and Intercepts of One Item Across Time

### Configural Invariance With Correlated Residuals Within-Indicator Across Time

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's factor loading to be the same across time
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. Allow within-indicator residuals to covary across time

```{r}
configuralInvarianceStandardizeT1Latent_syntax <- '
  # Factor Loadings
  latent_t1 =~ NA*load1*sad1 + load12*down1 + load13*depressed1
  latent_t2 =~ NA*load1*sad2 + load22*down2 + load23*depressed2
  latent_t3 =~ NA*load1*sad3 + load32*down3 + load33*depressed3
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  latent_t1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  latent_t1 ~~ 1*latent_t1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  latent_t2 ~ 1
  latent_t3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  latent_t2 ~~ latent_t2
  latent_t3 ~~ latent_t3
  
  # Fix Intercepts of Indicator 1 Across Time
  sad1 ~ inta*1
  sad2 ~ inta*1
  sad3 ~ inta*1
  
  # Free Intercepts of Remaining Manifest Variables
  down1 ~ intb1*1
  down2 ~ intb2*1
  down3 ~ intb3*1
  
  depressed1 ~ intc1*1
  depressed2 ~ intc2*1
  depressed3 ~ intc3*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
'

configuralInvarianceStandardizeT1Latent_fit <- cfa(
  configuralInvarianceStandardizeT1Latent_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)

summary(
  configuralInvarianceStandardizeT1Latent_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Metric ("Weak Factorial") Invariance

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. Allow within-indicator residuals to covary across time
1. **For each indicator, constrain its factor loading to be the same across time**

```{r}
metricInvarianceStandardizeT1Latent_syntax <- '
  # Factor Loadings
  latent_t1 =~ NA*load1*sad1 + load2*down1 + load3*depressed1
  latent_t2 =~ NA*load1*sad2 + load2*down2 + load3*depressed2
  latent_t3 =~ NA*load1*sad3 + load2*down3 + load3*depressed3
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  latent_t1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  latent_t1 ~~ 1*latent_t1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  latent_t2 ~ 1
  latent_t3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  latent_t2 ~~ latent_t2
  latent_t3 ~~ latent_t3
  
  # Fix Intercepts of Indicator 1 Across Time
  sad1 ~ inta*1
  sad2 ~ inta*1
  sad3 ~ inta*1
  
  # Free Intercepts of Remaining Manifest Variables
  down1 ~ intb1*1
  down2 ~ intb2*1
  down3 ~ intb3*1
  
  depressed1 ~ intc1*1
  depressed2 ~ intc2*1
  depressed3 ~ intc3*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
'

metricInvarianceStandardizeT1Latent_fit <- cfa(
  metricInvarianceStandardizeT1Latent_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)

summary(
  metricInvarianceStandardizeT1Latent_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

anova(
  configuralInvarianceStandardizeT1Latent_fit,
  metricInvarianceStandardizeT1Latent_fit
)
```

### Scalar ("Strong Factorial") Invariance

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time
1. For each indicator, constrain its factor loading to be the same across time
1. **For each indicator, constrain its intercept to be the same across time**

```{r}
scalarInvarianceStandardizeT1Latent_syntax <- '
  # Factor Loadings
  latent_t1 =~ NA*load1*sad1 + load2*down1 + load3*depressed1
  latent_t2 =~ NA*load1*sad2 + load2*down2 + load3*depressed2
  latent_t3 =~ NA*load1*sad3 + load2*down3 + load3*depressed3
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  latent_t1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  latent_t1 ~~ 1*latent_t1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  latent_t2 ~ 1
  latent_t3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  latent_t2 ~~ latent_t2
  latent_t3 ~~ latent_t3
  
  # Fix Intercepts Across Time
  sad1 ~ inta*1
  sad2 ~ inta*1
  sad3 ~ inta*1
  
  down1 ~ intb*1
  down2 ~ intb*1
  down3 ~ intb*1
  
  depressed1 ~ intc*1
  depressed2 ~ intc*1
  depressed3 ~ intc*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
'

scalarInvarianceStandardizeT1Latent_fit <- cfa(
  scalarInvarianceStandardizeT1Latent_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)

summary(
  scalarInvarianceStandardizeT1Latent_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

anova(
  scalarInvarianceStandardizeT1Latent_fit,
  metricInvarianceStandardizeT1Latent_fit
)
```

### Partial Strong Factorial Invariance

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time
1. For some indicators (i.e., the indicators with invariant factor loadings, called "anchor items"), constrain its factor loading to be the same across time
1. For some indicators (i.e., the indicators with invariant intercepts, called "anchor items"), constrain its intercept to be the same across time

```{r}
partialStrongInvarianceStandardizeT1Latent_syntax <- '
  # Factor Loadings
  latent_t1 =~ NA*load1*sad1 + load2*down1 + load3*depressed1
  latent_t2 =~ NA*load1*sad2 + load2*down2 + load3*depressed2
  latent_t3 =~ NA*load1*sad3 + load2*down3 + load33*depressed3
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  latent_t1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  latent_t1 ~~ 1*latent_t1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  latent_t2 ~ 1
  latent_t3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  latent_t2 ~~ latent_t2
  latent_t3 ~~ latent_t3
  
  # Fix Some Intercepts Across Time
  sad1 ~ inta*1
  sad2 ~ inta*1
  sad3 ~ inta*1
  
  down1 ~ intb*1
  down2 ~ intb2*1
  down3 ~ intb*1
  
  depressed1 ~ intc*1
  depressed2 ~ intc2*1
  depressed3 ~ intc*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
'

partialStrongInvarianceStandardizeT1Latent_fit <- cfa(
  partialStrongInvarianceStandardizeT1Latent_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)

summary(
  partialStrongInvarianceStandardizeT1Latent_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

anova(
  configuralInvarianceStandardizeT1Latent_fit,
  partialStrongInvarianceStandardizeT1Latent_fit
)
```

### Partial Strong Factorial Invariance With Second-Order Growth Curve {#sec-secondOrderLGCM}

#### Model Syntax

```{r}
partialStrongInvarianceGrowthCurve_syntax <- '
  # Factor Loadings
  latent_t1 =~ NA*load1*sad1 + load2*down1 + load3*depressed1
  latent_t2 =~ NA*load1*sad2 + load2*down2 + load3*depressed2
  latent_t3 =~ NA*load1*sad3 + load2*down3 + load33*depressed3
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  latent_t1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  latent_t1 ~~ 1*latent_t1
  
  # Fix Some Intercepts Across Time
  sad1 ~ inta*1
  sad2 ~ inta*1
  sad3 ~ inta*1
  
  down1 ~ intb*1
  down2 ~ intb2*1
  down3 ~ intb*1
  
  depressed1 ~ intc*1
  depressed2 ~ intc2*1
  depressed3 ~ intc*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
  
  # Linear Growth Model
  i =~ 1*latent_t1 + 1*latent_t2 + 1*latent_t3
  s =~ 0*latent_t1 + 1*latent_t2 + 2*latent_t3
  
  # Variance-covariances of intercepts and slopes
  i ~~ i
  s ~~ s
  i ~~ s
  
  # Means of level and slope
  i ~ 1
  s ~ 1
'
```

#### Fit Model

```{r}
partialStrongInvarianceGrowthCurve_fit <- cfa(
  partialStrongInvarianceGrowthCurve_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)
```

#### Model Summary

```{r}
summary(
  partialStrongInvarianceGrowthCurve_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

Parameter estimates:

```{r}
lavaan::parameterEstimates(partialStrongInvarianceGrowthCurve_fit) |>
  subset(
    lhs %in% c("i", "s") & substr(rhs, 1, 6) != "latent",
    -label
  )
```

# Alignment Optimization {#sec-alignment}

The example is adapted from the section on [longitudinal measurement invariance](#sec-longitudinalMI).
Two primary options for identifying a latent factor include:

1. standardizing the latent factor(s) at all timepoints (fixing their intercepts to zero and their variances to one)
1. standardizing the latent factor(s) at T1 (fixing their intercepts to zero and their variances to one) and constraining the factor loadings and intercepts of one item to be equal across time

(Option #1 may be preferable in alignment optimization.)

## Standardizing the Latent Factors at All Timepoints

### Configural Invariance With Correlated Residuals Within-Indicator Across Time

1. Standardize the latent factor(s) at all timepoints (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time

#### Model Syntax

```{r}
configuralInvarianceStandardizeAllLatent_syntax <- '
  # Factor Loadings
  latent_t1 =~ NA*load11*sad1 + load12*down1 + load13*depressed1
  latent_t2 =~ NA*load21*sad2 + load22*down2 + load23*depressed2
  latent_t3 =~ NA*load31*sad3 + load32*down3 + load33*depressed3
  
  # Factor Identification: Standardize Factors at All Timepoints (using std_lv = TRUE)
  
  # Free Intercepts of Manifest Variables
  sad1 ~ inta1*1
  sad2 ~ inta2*1
  sad3 ~ inta3*1
  
  down1 ~ intb1*1
  down2 ~ intb2*1
  down3 ~ intb3*1
  
  depressed1 ~ intc1*1
  depressed2 ~ intc2*1
  depressed3 ~ intc3*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
'
```

#### Fit Model

```{r}
configuralInvarianceStandardizeAllLatent_fit <- cfa(
  configuralInvarianceStandardizeAllLatent_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  std_lv = TRUE,
  fixed_x = FALSE)
```

#### Model Summary

```{r}
summary(
  configuralInvarianceStandardizeAllLatent_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Longitudinal Alignment Optimization

Adapted from Lai (2013; [code](https://github.com/marklhc/awc-growth-supp/blob/main/ex_MIDUS_growth.Rmd)):

```{r}
# Extract loadings and intercepts for alignment
lam_mat <- lavInspect(configuralInvarianceStandardizeAllLatent_fit, what = "est")$lambda
nu_vec <- lavInspect(configuralInvarianceStandardizeAllLatent_fit, what = "est")$nu

# Put them into T x p matrices
num_items <- 3
num_waves <- 3
lam_config <- crossprod(lam_mat, rep(1, num_waves) %x% diag(num_items))
nu_config <- matrix(nu_vec, nrow = num_waves, ncol = num_items, byrow = TRUE)

# Add indicator names
colnames(lam_config) <- colnames(nu_config) <- c("sad", "down", "depressed")

# Alignment optimization
aligned_pars <- sirt::invariance.alignment(
  lambda = lam_config,
  nu = nu_config,
  fixed = TRUE
)
```

The aligned loadings and intercepts are below:

```{r}
alignedParameters <- rbind(
  "Loadings" = rep(NA, num_items),
  `rownames<-`(aligned_pars$lambda.aligned,
               paste0("Time", 1:num_waves)),
  "Intercepts" = rep(NA, num_items),
  `rownames<-`(aligned_pars$nu.aligned,
               paste0("Time", 1:num_waves))
)

#opts <- options(knitr.kable.NA = "")
#
#knitr::kable(
#  alignedParameters,
#  format = "simple",
#  digits = 3L
#)

print(alignedParameters, na.print = "")

aligned_pars$lambda.aligned
aligned_pars$nu.aligned
```

#### Effect Size of Noninvariance ($d_{MACS}$)

Lai (2023) suggests using a "50/30/20" rule of thumb for using $d_{MACS}$ to evaluate the appropriateness of the alignment optimization method:
Alignment optimization within CFA "is trustworthy when (a) no more than 50% of items have one or more $d_{MACS}$ > .20 and (b) no more than 30% of the pairwise $d_{MACS}$ > .20.
Lai (2023) suggests reporting the range of $d_{MACS}$ values, the proportion of $d_{MACS}$ > .20, and the proportion of items with at least one $d_{MACS}$ > .20.

```{r}
# Function for dMACS
dmacs <- function(loadings, intercepts, pooled_item_sd,
                  latent_mean = 0, latent_var = 1) {
  dloading <- diff(loadings)
  dintercept <- diff(intercepts)
  integral <- dintercept^2 + 2 * dintercept * dloading * latent_mean +
    dloading^2 * (latent_var + latent_mean^2)
  sqrt(integral) / pooled_item_sd
}

# Use item SDs at first time point
item_sds_wave1 <-
  apply(
    mydata_wide[c("sad1", "down1", "depressed1")],
    2, sd
  )

dmacs_pairwise <- function(loading_mat, intercept_mat, pooled_item_sd,
                           latent_mean = 0, latent_var = 1) {
  ngroups <- nrow(loading_mat)
  pairs <- combn(ngroups, 2)
  out <- matrix(NA, nrow = ncol(pairs), ncol = ncol(loading_mat))
  for (i in seq_len(ncol(pairs))) {
    out[i, ] <- dmacs(loading_mat[pairs[, i], ],
      intercepts = intercept_mat[pairs[, i], ],
      pooled_item_sd,
      latent_mean,
      latent_var
    )
  }
  rownames(out) <- apply(pairs, 2, paste, collapse = " vs ")
  colnames(out) <- colnames(loading_mat)
  out
}

# All pairwise dMACS
dmacs_pairwise(aligned_pars$lambda.aligned,
  intercept_mat = aligned_pars$nu.aligned,
  pooled_item_sd = item_sds_wave1,
  latent_mean = 0,
  latent_var = 1
)
```

### An Equivalent Configural Model with Aligned Loadings

For the reference indicator (for which the loadings and intercepts are fixed), it is preferable to select an indicator with a strong factor loading (Lai, 2023).

```{r}
equivalentConfiguralModelAligned_syntax <- '
  # Factor Loadings (Loadings of first indicator fixed to alignment solution)
  latent_t1 =~ 0.9904750*load11*sad1 + load12*down1 + load13*depressed1
  latent_t2 =~ 0.9880715*load21*sad2 + load22*down2 + load23*depressed2
  latent_t3 =~ 0.9863225*load31*sad3 + load32*down3 + load33*depressed3
  
  # Intercepts of Manifest Variables (Intercepts of first indicator fixed to alignment solution)
  sad1 ~ 3.319192*inta1*1
  sad2 ~ 3.332276*inta2*1
  sad3 ~ 3.345631*inta3*1
  
  down1 ~ intb1*1
  down2 ~ intb2*1
  down3 ~ intb3*1
  
  depressed1 ~ intc1*1
  depressed2 ~ intc2*1
  depressed3 ~ intc3*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
  
  # Free latent means
  latent_t1 + latent_t2 + latent_t3 ~ NA*1
'

equivalentConfiguralModelAligned_fit <- cfa(
  equivalentConfiguralModelAligned_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)
```

Compare model fit (they're equivalent):

```{r}
anova(
  configuralInvarianceStandardizeAllLatent_fit,
  equivalentConfiguralModelAligned_fit
)
```

### Alignment-Within-CFA (AwC) Approach for Growth Modeling

#### Model Syntax

For the reference indicator (for which the loadings and intercepts are fixed), it is preferable to select an indicator with a strong factor loading (Lai, 2023).

```{r}
awcGrowthModel_syntax <- '
  # Factor Loadings (Loadings of first indicator fixed to alignment solution)
  latent_t1 =~ 0.9904750*load11*sad1 + load12*down1 + load13*depressed1
  latent_t2 =~ 0.9880715*load21*sad2 + load22*down2 + load23*depressed2
  latent_t3 =~ 0.9863225*load31*sad3 + load32*down3 + load33*depressed3
  
  # Intercepts of Manifest Variables (Intercepts of first indicator fixed to alignment solution)
  sad1 ~ 3.319192*inta1*1
  sad2 ~ 3.332276*inta2*1
  sad3 ~ 3.345631*inta3*1
  
  down1 ~ intb1*1
  down2 ~ intb2*1
  down3 ~ intb3*1
  
  depressed1 ~ intc1*1
  depressed2 ~ intc2*1
  depressed3 ~ intc3*1
  
  # Residual Covariances Within Indicator Across Time
  sad1 ~~ sad2
  sad2 ~~ sad3
  sad3 ~~ sad3
  
  down1 ~~ depressed2
  down2 ~~ depressed3
  down3 ~~ depressed3
  
  depressed1 ~~ depressed2
  depressed2 ~~ depressed3
  depressed3 ~~ depressed3
  
  # Linear Growth Model
  i =~ 1*latent_t1 + 1*latent_t2 + 1*latent_t3
  s =~ 0*latent_t1 + 1*latent_t2 + 2*latent_t3
  
  # Variance-covariances of intercepts and slopes
  i ~~ i
  s ~~ s
  i ~~ s
  
  # Means of level and slope
  i ~ 1
  s ~ 1
  
  # Fixed disturbances of latent outcomes to zero
  latent_t1 ~ 0*1
  latent_t2 ~ 0*1
  latent_t3 ~ 0*1
'
```

#### Fit Model

```{r}
awcGrowthModel_fit <- cfa(
  awcGrowthModel_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std_lv = TRUE,
  fixed_x = FALSE)
```

#### Model Summary

```{r}
summary(
  configuralInvarianceStandardizeAllLatent_fit,
  fit_measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

Parameter estimates:

```{r}
lavaan::parameterEstimates(awcGrowthModel_fit) |>
  subset(
    lhs %in% c("i", "s") & substr(rhs, 1, 6) != "latent",
    -label
  ) |>
  knitr::kable(format = "simple", digits = 3L)
```

## Standardizing the Latent Factors at T1 while Constraining Loadings and Intercepts of One Item Across Time

### Configural Invariance With Correlated Residuals Within-Indicator Across Time

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's factor loading to be the same across time
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. **Allow within-indicator residuals to covary across time**

# Moderated Nonlinear Factor Analysis (MNLFA) {#sec-mnlfa}

# Item Response Theory (IRT) with Anchor Items {#sec-irt}

It can be potentially helpful to consider regularization/Bayesian penalty methods (e.g., using priors such as Horseshoe, Lasso, or Spike-and-slab).

For examples of IRT models, see [here](irt.qmd).

## Separate Calibration {#sec-irtSeparateCalibration}

### Frequentist

```{r}
#| output: false

fit1 <- mirt::mirt(dat1[item_names_T1], 1, itemtype = "graded", SE = TRUE)
fit2 <- mirt::mirt(dat2[item_names_T2], 1, itemtype = "graded", SE = TRUE)
fit3 <- mirt::mirt(dat3[item_names_T3], 1, itemtype = "graded", SE = TRUE)
```

```{r}
# EAP theta estimates on each timepoint's OWN (arbitrary N(0,1)) metric
th1_sep <- mirt::fscores(fit1, method = "EAP")
th2_sep <- mirt::fscores(fit2, method = "EAP")
th3_sep <- mirt::fscores(fit3, method = "EAP")

# pull item parameters in plink's expected format: a, b1, b2, b3
cfs1 <- mirt::coef(fit1, IRTpars = TRUE, simplify = TRUE)$items[, c("a", "b1", "b2", "b3")]
cfs2 <- mirt::coef(fit2, IRTpars = TRUE, simplify = TRUE)$items[, c("a", "b1", "b2", "b3")]
cfs3 <- mirt::coef(fit3, IRTpars = TRUE, simplify = TRUE)$items[, c("a", "b1", "b2", "b3")]
rownames(cfs1) <- item_names_T1
rownames(cfs2) <- item_names_T2
rownames(cfs3) <- item_names_T3

# build the plink irt.pars object
common_names_12 <- paste0("C12_", 1:n_common)
common_names_23 <- paste0("C23_", 1:n_common)

common12 <- cbind(
  match(common_names_12, rownames(cfs1)),
  match(common_names_12, rownames(cfs2)))
common23 <- cbind(
  match(common_names_23, rownames(cfs2)),
  match(common_names_23, rownames(cfs3)))

pm1 <- plink::as.poly.mod(nrow(cfs1), "grm", 1:nrow(cfs1))
pm2 <- plink::as.poly.mod(nrow(cfs2), "grm", 1:nrow(cfs2))
pm3 <- plink::as.poly.mod(nrow(cfs3), "grm", 1:nrow(cfs3))

plinkIRTpars <- plink::as.irt.pars(
  x        = list(cfs1, cfs2, cfs3),
  common   = list(common12, common23),
  cat      = list(rep(n_cat, nrow(cfs1)), rep(n_cat, nrow(cfs2)), rep(n_cat, nrow(cfs3))),
  poly.mod = list(pm1, pm2, pm3),
  grp.names = c("T1", "T2", "T3")
)

# Stocking-Lord chain linking, T1 <-> T2 <-> T3, rescaled onto T2 (base.grp = 2)
link_out <- plink::plink(
  plinkIRTpars,
  rescale = "SL",
  base.grp = 2,
  ability = list(th1_sep, th2_sep, th3_sep)
)

summary(link_out, descrip = TRUE)             # linking constants + common-item drift diagnostics
constants <- plink::link.con(link_out)        # A, B for each transformation
theta_linked <- plink::link.ability(link_out) # list: theta1, theta2, theta3 all on the T2 metric

th1_linked <- theta_linked[[1]]
th2_linked <- theta_linked[[2]]          # unchanged (T2 is the base)
th3_linked <- theta_linked[[3]]
```

### Bayesian

## Concurrent Calibration {#sec-irtConcurrentCalibration}

### Frequentist

```{r}
# Anchor items are constrained equal across the groups that share them;
# unique items are freely (and separately) estimated group-by-group because
# they are 100% missing outside their own timepoint and so contribute nothing
# to the other groups' likelihood. Latent means/variances are freed for all
# but the reference group (T1), which is fixed at N(0,1) for identification.

anchor_items <- union(common_names_12, common_names_23)
```

```{r}
#| output: false

fit_concurrent <- mirt::multipleGroup(
  dat_concurrent[all_items],
  model = 1,
  group = dat_concurrent$timepoint,
  itemtype = "graded",
  invariance = c(anchor_items, "free_means", "free_var"),
  SE = TRUE
)
```

```{r}
# growth in the latent trait, directly comparable across timepoints
mirt::coef(fit_concurrent, simplify = TRUE) # inspect per-group means/variances

theta_concurrent <- mirt::fscores(fit_concurrent, method = "EAP")
th1_conc <- theta_concurrent[dat_concurrent$timepoint == "T1"]
th2_conc <- theta_concurrent[dat_concurrent$timepoint == "T2"]
th3_conc <- theta_concurrent[dat_concurrent$timepoint == "T3"]
```

Compare results from separate vs concurrent calibration against truth:

```{r}
cat("correlation with true theta (T1): separate calibration =",
    round(cor(th1_linked, theta1_true), 3),
    " | concurrent calibration =", round(cor(th1_conc, theta1_true), 3), "\n")
cat("correlation with true theta (T3): separate calibration =",
    round(cor(th3_linked, theta3_true), 3),
    " | concurrent calibration =", round(cor(th3_conc, theta3_true), 3), "\n")

cat("recovered mean growth T1->T3 (true = 1.0):\n",
    " separate calibration:", round(mean(th3_linked) - mean(th1_linked), 3), "\n",
    " concurrent calibration:", round(mean(th3_conc) - mean(th1_conc),  3))
```

### Bayesian

```{r}
formula_grm <- brms::bf(
  response ~ beta + exp(alpha) * theta,
  theta ~ 0 + timeCentered + (1 + timeCentered | idFactor),   # person parameter
  beta ~ 1 + timeCentered + (1 + timeCentered | itemFactor),  # item parameter: easiness
  alpha ~ 1 + timeCentered + (1 + timeCentered | itemFactor), # item parameter: discrimination
  nl = TRUE
)

modelPriors <- c(
  prior(constant(1), class = "sd", group = "idFactor", nlpar = "theta")
)

# Parallel Processing
rstan_options(auto_write = TRUE)

fallback <- parallel::detectCores()
NSLOTS <- as.integer(Sys.getenv("NSLOTS"))
if(is.na(NSLOTS)) NSLOTS <- fallback
options(mc.cores = NSLOTS)

print("Slots:")
print(as.integer(Sys.getenv("NSLOTS")))

print("Cores:")
print(parallel::detectCores())
print(parallelly::availableCores())
```

```{r}
#| eval: false

fit_grm <- brms::brm(
  formula = formula_grm,
  data = dat_long,
  family = brmsfamily("cumulative", "logit"),
  prior = modelPriors,
  iter = 2000, # takes a long time to fit
  cores = 4,
  threads = threading(parallelly::availableCores()),
  seed = 52242,
  silent = 0
)
```

```{r}
#| include: false
#| eval: false

save(
  fit_grm,
  "fit_grm.rdata"
)
```

```{r}
#| include: false
#| eval: false

load("fit_grm.rdata")
```

```{r}
#| eval: false

summary(fit_grm)

print(effectsize::standardize_parameters(
  fit_grm,
  method = "posthoc"),
  digits = 2)

performance::model_performance(fit_grm)
performance::r2(fit_grm)
```
