---
title: "Exploratory Data Analysis"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      comment = "")
```

# Rationale

Exploratory data analysis is important for understanding your data, checking for data issues/errors, and checking assumptions for different statistical models.

LOOK AT YOUR DATA—this is one of the most overlooked steps in data analysis!

# Preamble

## Install Libraries

```{r}
#install.packages("remotes")
#remotes::install_git("https://research-git.uiowa.edu/PetersenLab/petersenlab.git")
```

## Load Libraries

```{r, message = FALSE, warning = FALSE}
library("petersenlab")
library("car")
library("vioplot")
library("ellipse")
library("nlme")
library("effects")
library("corrplot")
library("ggplot2")
library("psych")
library("tidyverse")
library("purrr")
library("naniar")
library("mvnormtest")
```

# Simulate Data

```{r}
set.seed(52242)

n <- 1000

ID <- rep(1:100, each = 10)
predictor <- rbeta(n, 1.5, 5) * 100
outcome <- predictor + rnorm(n, mean = 0, sd = 20) + 50

predictorOverplot <- sample(1:50, n, replace = TRUE)
outcomeOverplot <- predictorOverplot + sample(1:75, n, replace = TRUE)

categorical1 <- sample(1:5, size = n, replace = TRUE)
categorical2 <- sample(1:5, size = n, replace = TRUE)

mydata <- data.frame(ID = ID,
                     predictor = predictor,
                     outcome = outcome,
                     predictorOverplot = predictorOverplot,
                     outcomeOverplot = outcomeOverplot,
                     categorical1 = categorical1,
                     categorical2 = categorical2)

mydata[sample(1:n, size = 10), "predictor"] <- NA
mydata[sample(1:n, size = 10), "outcome"] <- NA
mydata[sample(1:n, size = 10), "predictorOverplot"] <- NA
mydata[sample(1:n, size = 10), "outcomeOverplot"] <- NA
mydata[sample(1:n, size = 30), "categorical1"] <- NA
mydata[sample(1:n, size = 70), "categorical2"] <- NA
```

# Descriptive statistics

```{r}
round(data.frame(psych::describe(mydata)), 2)
```

## Sample

- Check the sample size (*N*)
  - Is the sample size in the data the expected sample size?
  Are there cases (participants) that are missing?
  Are there cases that should not be there?
  - Here is the sample size:
  
```{r}
length(unique(mydata$ID))
```   

- Check the extent of missingness
    - How much data are missing in the model variables—including the predictor, outcome, and covariates?
    - Here are the proportion of missing data in each variable:
    
```{r}
map(mydata, ~mean(is.na(.))) %>% t %>% t
```

## Distribution

- Frequencies
    - Examine the frequencies of categorical variables:

```{r}
mydata %>% 
  select(categorical1, categorical2) %>%
  sapply(function(x) table(x, useNA = "always")) %>% 
  t()
```

## Central Tendency

- Mean

```{r}
round(colMeans(mydata, na.rm = TRUE), 2)
round(apply(mydata, 2, function(x) mean(x, na.rm = TRUE)), 2)

mydata %>% 
  summarise_all(list(~ mean(., na.rm = TRUE))) %>% 
  round(., 2)
```

- Median

```{r}
round(apply(mydata, 2, function(x) median(x, na.rm = TRUE)), 2)

mydata %>% 
  summarise_all(list(~ median(., na.rm = TRUE))) %>% 
  round(., 2)
```

- Mode

```{r}
round(apply(mydata, 2, function(x) Mode(x, multipleModes = "mean")), 2)

mydata %>% 
  summarise_all(list(~ Mode(., multipleModes = "mean"))) %>% 
  round(., 2)
```

Compute all of these measures of central tendency:

```{r}
mydata %>% 
  summarise_all(list(~ mean(., na.rm = TRUE),
                     ~ median(., na.rm = TRUE),
                     ~ Mode(., multipleModes = "mean"))) %>% 
  round(., 2) %>% 
  pivot_longer(cols = everything(),
               names_to = c("variable","index"),
               names_sep = "_") %>% 
  pivot_wider(names_from = index,
              values_from = value) %>% 
  rename(mode = Mode)
```

## Dispersion

- Standard deviation
- Observed minimum and maximum (vis-à-vis possible minimum and maximum)
- Skewness
- Kurtosis

Compute all of these measures of dispersion:

```{r}
mydata %>% 
  summarise_all(list(~ sd(., na.rm = TRUE),
                     ~ min(., na.rm = TRUE),
                     ~ max(., na.rm = TRUE),
                     ~ skew(., na.rm = TRUE),
                     ~ kurtosi(., na.rm = TRUE))) %>% 
  round(., 2) %>% 
  pivot_longer(cols = everything(),
               names_to = c("variable","index"),
               names_sep = "_") %>% 
  pivot_wider(names_from = index,
              values_from = value) %>% 
  rename(skewness = skew,
         kurtosis = kurtosi)
```

Consider transforming data if skewness > |0.8| or if kurtosis > |3.0|.

## Summary Statistics

Add summary statistics to the bottom of correlation matrices in papers:

```{r}
summaryTable <- mydata %>% 
  summarise_all(list(~ length(na.omit(.)),
                     ~ mean(is.na(.)) * 100,
                     ~ mean(., na.rm = TRUE),
                     ~ sd(., na.rm = TRUE),
                     ~ skew(., na.rm = TRUE),
                     ~ kurtosi(., na.rm = TRUE))) %>% 
  round(., 2) %>% 
  pivot_longer(cols = everything(),
               names_to = c("variable","index"),
               names_sep = "_") %>% 
  pivot_wider(names_from = index,
              values_from = value) %>% 
  rename(n = length,
         missingness = "*",
         M = mean,
         SD = sd,
         skewness = skew,
         kurtosis = kurtosi)

summaryTableTransposed <- summaryTable[-1] %>% 
  t() %>% 
  as.data.frame() %>% 
  setNames(summaryTable$variable) %>% 
  round(., digits = 2)

cor.table(mydata, type = "manuscript")
summaryTableTransposed
```

## Distribution Plots

See [here](figures.html) for resources for creating figures in R.

### Histogram

#### Base R

```{r}
hist(mydata$outcome)
```

#### `ggplot2`

```{r}
ggplot(mydata, aes(x = outcome)) +
  geom_histogram(color = 1)
```

### Histogram overlaid with density plot and rug plot

#### Base R

```{r}
hist(mydata$outcome, prob = TRUE)
lines(density(mydata$outcome, na.rm = TRUE))
rug(mydata$outcome)
```

#### `ggplot2`

```{r}
ggplot(mydata, aes(x = outcome)) +
  geom_histogram(aes(y = ..density..), color = 1) +
  geom_density() +
  geom_rug()
```

### Density Plot

#### Base R

```{r}
plot(density(mydata$outcome, na.rm = TRUE))
```

#### `ggplot2`

```{r}
ggplot(mydata, aes(x = outcome)) +
  geom_density()
```

### Box and whisker plot (boxplot)

#### Base R

```{r}
boxplot(mydata$outcome, horizontal = TRUE)
```

#### `ggplot2`

```{r}
ggplot(mydata, aes(x = outcome)) +
  geom_boxplot()
```

### Violin plot

#### Base R

```{r}
vioplot(na.omit(mydata$outcome), horizontal = TRUE)
```

#### `ggplot2`

```{r}
ggplot(mydata, aes(x = "", y = outcome)) +
  geom_violin()
```

# Bivariate Associations

## High Density Scatterplot

```{r}
ggplot(mydata, aes(x = predictorOverplot, y = outcomeOverplot)) + 
  geom_point(position = "jitter", alpha = 0.3) + 
  geom_density2d()

smoothScatter(mydata$predictorOverplot, mydata$outcomeOverplot)
```

## Data Ellipse

```{r}
mydata_nomissing <- na.omit(mydata[,c("predictor","outcome")])
dataEllipse(mydata_nomissing$predictor, mydata_nomissing$outcome, levels = c(0.5, .95))
```

## Visually Weighted Regression

```{r, message = FALSE, results = "hide"}
vwReg(outcome ~ predictor, data = mydata)
```

# Basic inferential statistics

## Tests of Normality

### Shapiro-Wilk test of normality

The Shapiro-Wilk test of normality does not accept more than 5000 cases because it will reject the hypothesis that data come from a normal distribution with even slight deviations from normality.

```{r}
shapiro.test(na.omit(mydata$outcome)) #subset to keep only the first 5000 rows: mydata$outcome[1:5000]
```

### Test of multivariate normality

```{r}
mydata %>% 
  na.omit %>% 
  t %>% 
  mshapiro.test
```

## Statistical decision tree

## Tests of systematic missingness (i.e., whether missingness on a variable depends on other variables)

- Generally test:
    - Whether data are consistent with a missing completely at random (MCAR) pattern—Little's MCAR Test
    - Whether outcome variable(s) differ as a function of any model variables (predictors and covariates) and as a function of any key demographic characteristics (e.g., sex, ethnicity, socioeconomic status)
    - Whether focal predictor variable(s) differ as a function of any model variables (including outcome variable) and as a function of any key demographic characteristics
- For instance:
    - Whether males are more likely than girls to be missing scores on the dependent variable
    - Whether longitudinal attrition is greater in lower socioeconomic status families
- If missingness differs systematically as a function of other variables, you can include that variable as a control variable in models, and/or can include that variable in multiple imputation to inform imputed scores for missing values

### Little's MCAR Test

```{r}
mcar_test(mydata)
```

## Multivariate Associations

### Correlation Matrix

#### Pearson Correlations

```{r}
cor.table(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")])
cor.table(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")], type = "manuscript")
cor.table(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")], type = "manuscriptBig")
```

#### Spearman Correlations

```{r}
cor.table(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")], correlation = "spearman")
cor.table(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")], type = "manuscript", correlation = "spearman")
cor.table(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")], type = "manuscriptBig", correlation = "spearman")
```

#### Partial Correlations

Examine the associations among variables controlling for a covariate (`outcomeOverplot`).

```{r}
partialcor.table(mydata[,c("predictor","outcome","predictorOverplot")], z = mydata[,c("outcomeOverplot")])
partialcor.table(mydata[,c("predictor","outcome","predictorOverplot")], z = mydata[,c("outcomeOverplot")], type = "manuscript")
partialcor.table(mydata[,c("predictor","outcome","predictorOverplot")], z = mydata[,c("outcomeOverplot")], type = "manuscriptBig")
```

### Correlogram

```{r}
corrplot(cor(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")], use = "pairwise.complete.obs"))
```

### Scatterplot matrix

```{r}
scatterplotMatrix(~ predictor + outcome + predictorOverplot + outcomeOverplot, data = mydata, use = "pairwise.complete.obs")
```

### Pairs panels

```{r}
pairs.panels(mydata[,c("predictor","outcome","predictorOverplot","outcomeOverplot")])
```

## Effect Plots

### Multiple Regression Model

```{r}
multipleRegressionModel <- lm(outcome ~ predictor + predictorOverplot,
                              data = mydata,
                              na.action = "na.exclude")

allEffects(multipleRegressionModel)
plot(allEffects(multipleRegressionModel))
```

### Multilevel Regression Model

```{r}
multilevelRegressionModel <- lme(outcome ~ predictor + predictorOverplot, random = ~ 1|ID,
                                 method = "ML",
                                 data = mydata,
                                 na.action = "na.exclude")

allEffects(multilevelRegressionModel)
plot(allEffects(multilevelRegressionModel))
```

# Session Info

```{r, class.source = "fold-hide"}
sessionInfo()
```
