---
title: "Figures in R"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  warning = FALSE,
  comment = "")
```

# Picking a Chart Type

https://www.data-to-viz.com

# Gallery

https://www.r-graph-gallery.com

# Resources for Learning `R` Syntax for Figures

https://www.statmethods.net/graphs/index.html

http://www.cookbook-r.com/Graphs/

# Petersen Lab Examples

https://research-git.uiowa.edu/PetersenLab/R-Plotting/-/tree/main/Analyses

# Font

You can download the lab fonts for figures here[^1]: https://drive.google.com/drive/u/0/folders/1fqlrnEe7NFnWZoIrsHmr8ulDS4nhs-H3

## Preamble

### Install Libraries

```{r}
#install.packages("remotes")
#remotes::install_github("DevPsyLab/petersenlab")
```

### Load Libraries

```{r, message = FALSE, warning = FALSE}
library("petersenlab")
library("ellipse")
library("ggplot2")
library("grid")
library("reshape")
library("plyr")
library("RColorBrewer")
library("reshape2")
library("ggExtra")
library("viridis")
library("ggthemes")
library("ggpubr")
```

## Simulate Data

```{r}
set.seed(52242)

n <- 1000

predictor <- rbeta(n, 1.5, 5) * 100
outcome <- predictor + rnorm(n, mean = 0, sd = 20) + 50
number <- sample(1:1000, replace = TRUE)

predictorOverplot <- sample(1:50, n, replace = TRUE)
outcomeOverplot <- predictorOverplot + sample(1:75, n, replace = TRUE)

df <- data.frame(predictor = predictor,
                 outcome = outcome,
                 predictorOverplot = predictorOverplot,
                 outcomeOverplot = outcomeOverplot)

df[sample(1:n, size = 10), "predictor"] <- NA
df[sample(1:n, size = 10), "outcome"] <- NA
df[sample(1:n, size = 10), "predictorOverplot"] <- NA
df[sample(1:n, size = 10), "outcomeOverplot"] <- NA
```

## Line

```{r}
plot.new()
lines(
  x = seq(from = -10, to = 10, length.out = 100),
  y = seq(from = -25, to = 25, length.out = 100))
```

## Curve

```{r}
curve(x^3 - 3*x, from = -2, to = 2)
curve(x^2 - 2, add = TRUE, col = "violet")
```

## Basic Scatterplot

### Base R

```{r}
plot(outcome ~ predictor, data = df)
plot(df$predictor, df$outcome)
```

#### Best-fit line

```{r}
plot(outcome ~ predictor, data = df)
abline(lm(outcome ~ predictor, data = df), col = "red") #regression line (y~x)
```

#### Best-fit line with correlation coefficient

```{r}
plot(outcome ~ predictor, data = df)
abline(lm(outcome ~ predictor, data = df), col = "red") #regression line (y~x)
addText(x = df$predictor, y = df$outcome)
```

#### Loess line

```{r}
plot(outcome ~ predictor, data = df)
lines(loess.smooth(df$predictor, df$outcome)) #loess line (x,y)
```

### `ggplot2`

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  theme_classic()
```

#### Best-fit line

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  theme_classic()
```

#### Best-fit line with correlation coefficient

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  stat_cor(
    cor.coef.name = "r",
    p.accuracy = 0.001,
    r.accuracy = 0.01) +
  theme_classic()
```

#### Loess line

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  stat_smooth(method = "loess", formula = y ~ x) +
  theme_classic()
```

## Change Plot Style

### Change Theme

```{r}
basePlot <- ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point()
```

#### Default Theme

```{r}
basePlot
```

#### Grayscale: `theme_gray()`

```{r}
basePlot + theme_gray() + theme(text = element_text(family = "Gotham"))
```

#### Black-and-White: `theme_bw()`

```{r}
basePlot + theme_bw() + theme(text = element_text(family = "Gotham"))
```

#### Line Drawing: `theme_linedraw()`

A theme with only black lines of various widths on white backgrounds, reminiscent of a line drawing.
Note that this theme has some very thin lines (<< 1 pt) which some journals may refuse.

```{r}
basePlot + theme_linedraw() + theme(text = element_text(family = "Gotham"))
```

#### Light: `theme_light()`

```{r}
basePlot + theme_light() + theme(text = element_text(family = "Gotham"))
```

#### Dark: `theme_dark()`

```{r}
basePlot + theme_dark() + theme(text = element_text(family = "Gotham"))
```

#### Minimal: `theme_minimal()`

```{r}
basePlot + theme_minimal() + theme(text = element_text(family = "Gotham"))
```

#### Classic: `theme_classic()`

```{r}
basePlot + theme_classic() + theme(text = element_text(family = "Gotham"))
```

#### A Completely Empty Theme: `theme_void()`

```{r}
basePlot + theme_void() + theme(text = element_text(family = "Gotham"))
```

#### Visual Unit Tests: `theme_test()`

```{r}
basePlot + theme_test() + theme(text = element_text(family = "Gotham"))
```

#### Edward Tufte: `theme_tufte()`

Theme based on Edward Tufte.

```{r}
basePlot + theme_tufte()
```

#### Wall Street Journal: `theme_wsj()`

Theme based on the publication, the Wall Street Journal.

```{r}
basePlot + theme_wsj()
```

#### FiveThirtyEight: `theme_fivethirtyeight()`

Theme based on the publication, FiveThirtyEight.

```{r}
basePlot + theme_fivethirtyeight()
```

#### The Economist: `theme_economist()`

Theme based on the publication, The Economist.

```{r}
basePlot + theme_economist()
```

#### Stephen Few: `theme_few()`

Theme based on the rules and examples from Stephen Few's *Show Me the Numbers* and "Practical Rules for Using Color in Charts".

```{r}
basePlot + theme_few()
```

## Add Marginal Distributions {#marginalDistributions}

```{r}
scatterplot <- ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

### Density Plot

```{r}
densityMarginal <- ggMarginal(scatterplot, type = "density", xparams = list(fill = "gray"), yparams = list(fill = "gray"))
```

```{r}
print(densityMarginal, newpage = TRUE)
```

### Histogram

```{r}
histogramMarginal <- ggMarginal(scatterplot, type = "histogram", xparams = list(fill = "gray"), yparams = list(fill = "gray"))
```

```{r}
print(histogramMarginal, newpage = TRUE)
```

### Boxplot

```{r}
boxplotMarginal <- ggMarginal(scatterplot, type = "boxplot", xparams = list(fill = "gray"), yparams = list(fill = "gray"))
```

```{r}
print(boxplotMarginal, newpage = TRUE)
```

### Violin Plot

```{r}
violinMarginal <- ggMarginal(scatterplot, type = "violin", xparams = list(fill = "gray"), yparams = list(fill = "gray"))
```

```{r}
print(violinMarginal, newpage = TRUE)
```

### Density Plot and Histogram

```{r}
densigramMarginal <- ggMarginal(scatterplot, type = "densigram", xparams = list(fill = "gray"), yparams = list(fill = "gray"))
```

```{r}
print(densigramMarginal, newpage = TRUE)
```

## Ellipse

### Basic Ellipse

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  stat_ellipse(alpha = 0.4, level = 0.95, geom = "polygon", fill = "red", color = "red") +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

### Align Coordinates

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point() +
  stat_ellipse(alpha = 0.4, level = 0.95, geom = "polygon", fill = "red", color = "red") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_fixed(ratio = (max(predictor, na.rm = TRUE) - min(predictor, na.rm = TRUE))/(max(outcome, na.rm = TRUE) - min(outcome, na.rm = TRUE)),
              xlim = c(0, max(predictor, na.rm = TRUE)), 
              ylim = c(0, max(outcome, na.rm = TRUE))) +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

### Reduce Dot Size

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point(size = 0.5) +
  stat_ellipse(alpha = 0.4, level = 0.95, geom = "polygon", fill = "red", color = "red") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_fixed(ratio = (max(predictor, na.rm = TRUE) - min(predictor, na.rm = TRUE))/(max(outcome, na.rm = TRUE) - min(outcome, na.rm = TRUE)),
              xlim = c(0, max(predictor, na.rm = TRUE)), 
              ylim = c(0, max(outcome, na.rm = TRUE))) +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

### Transparency

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point(alpha = 0.3) +
  stat_ellipse(alpha = 0.4, level = 0.95, geom = "polygon", fill = "red", color = "red") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_fixed(ratio = (max(predictor, na.rm = TRUE) - min(predictor, na.rm = TRUE))/(max(outcome, na.rm = TRUE) - min(outcome, na.rm = TRUE)),
              xlim = c(0, max(predictor, na.rm = TRUE)), 
              ylim = c(0, max(outcome, na.rm = TRUE))) +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

## Bubble Chart

### Basic Bubble Chart

```{r}
ggplot(df, aes(x = predictorOverplot, y = outcomeOverplot)) +
  geom_count(aes(size = ..n..)) +
  scale_size_area() +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

### Specify Sizes

```{r}
ggplot(df, aes(x = predictorOverplot, y = outcomeOverplot)) +
  geom_count(aes(size = ..n..)) +
  scale_size_continuous(breaks = c(1, 2, 3, 4), range = c(1, 7)) +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

## 2-Dimensional Density

```{r}
ggplot(df, aes(x = predictor, y = outcome)) +
  stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_viridis() +
  theme(
    legend.position = "none",
    text = element_text(family = "Gotham")
  )
```

## Combined Ellipse and Bubble Chart

### `ggplot2`

```{r}
ggplot(df, aes(x = predictorOverplot, y = outcomeOverplot)) +
  geom_count(alpha = .6, color = rgb(0,0,.7,.5)) +
  scale_size_continuous(breaks = c(1, 2, 3, 4), range = c(1, 7)) +
  stat_smooth(method = "loess", se = TRUE, color = "green") + 
  stat_smooth(method = "lm") +
  stat_ellipse(alpha = 0.4, level = 0.95, geom = "polygon", fill = "red", color = "red") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_fixed(ratio = (max(predictorOverplot, na.rm = TRUE) - min(predictorOverplot, na.rm = TRUE))/(max(outcomeOverplot, na.rm = TRUE) - min(outcomeOverplot, na.rm = TRUE)),
              xlim = c(0, max(predictorOverplot, na.rm = TRUE)), 
              ylim = c(0, max(outcomeOverplot, na.rm = TRUE))) +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))

ggplot(df, aes(x = predictorOverplot, y = outcomeOverplot)) +
  geom_count(alpha = .6, color = rgb(0,0,.7,.5)) +
  scale_size_continuous(breaks = c(1, 2, 3, 4), range = c(1, 7)) +
  stat_smooth(method = "loess", se = TRUE, color = "green") + 
  stat_smooth(method = "lm") +
  stat_ellipse(color = "red", size = 1.5) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_fixed(ratio = (max(predictorOverplot, na.rm = TRUE) - min(predictorOverplot, na.rm = TRUE))/(max(outcomeOverplot, na.rm = TRUE) - min(outcomeOverplot, na.rm = TRUE)),
              xlim = c(0, max(predictorOverplot, na.rm = TRUE)), 
              ylim = c(0, max(outcomeOverplot, na.rm = TRUE))) +
  theme_classic() +
  theme(text = element_text(family = "Gotham"))
```

### Other implementation

From: https://stats.stackexchange.com/questions/7899/complex-regression-plot-in-r

#### `ggplot2`

```{r}
df$x <- df$predictorOverplot
df$y <- df$outcomeOverplot

xc <- with(df, xyTable(x, y))
df2 <- cbind.data.frame(x = xc$x, y = xc$y, number = xc$number)
df2$n <- cut(df2$number, c(0,1.5,2.5,Inf), labels = c(1,2,4))
df.ell <- as.data.frame(with(df, ellipse(cor(df$x, df$y, use = "pairwise.complete.obs"),
                                         scale = c(sd(df$x, na.rm = TRUE), sd(df$y, na.rm = TRUE)),
                                         centre = c(mean(df$x, na.rm = TRUE), mean(df$y, na.rm = TRUE)),
                                         level = .95)))

ggplot(data = na.omit(df2), aes(x = x, y = y)) + 
  geom_point(aes(size = n), alpha = .6, color = rgb(0,0,.7,.5)) + 
  stat_smooth(data = df, method = "loess", se = FALSE, color = "green") + 
  stat_smooth(data = df, method = "lm", col = "red") +
  geom_path(data = df.ell, colour = "green", size = 1) +
  coord_cartesian(xlim = c(-1,60), ylim = c(-1,130))
```

#### Base `R`

```{r}
do.it <- function(df, type="confidence", ...) {
  require(ellipse)
  lm0 <- lm(y ~ x, data=df)
  xc <- with(df, xyTable(x, y))
  df.new <- data.frame(x = seq(min(df$x), max(df$x), 0.1))
  pred.ulb <- predict(lm0, df.new, interval = type)
  pred.lo <- predict(loess(y ~ x, data = df), df.new)
  plot(xc$x, xc$y, cex = xc$number*1/4, xlab = "x", ylab = "y", ...) #change number*X to change dot size
  abline(lm0, col = "red")
  lines(df.new$x, pred.lo, col="green", lwd = 2)
  lines(df.new$x, pred.ulb[,"lwr"], lty = 2, col = "red")
  lines(df.new$x, pred.ulb[,"upr"], lty = 2, col = "red")    
  lines(ellipse(cor(df$x, df$y), scale=c(sd(df$x),sd(df$y)), 
                centre = c(mean(df$x), mean(df$y)), level = .95), lwd = 2, col = "green")
  invisible(lm0)
}

df3 <- na.omit(df[sample(nrow(df), nrow(df), rep = TRUE),])
df3$x <- df3$predictorOverplot
df3$y <- df3$outcomeOverplot

do.it(df3, pch = 19, col = rgb(0,0,.7,.5))
```

## Visually-Weighted Regression

### Default

```{r, message = FALSE, results = "hide"}
vwReg(outcome ~ predictor, data = df)
```

### Shade

```{r, message = FALSE, results = "hide"}
vwReg(outcome ~ predictor, data = df, shade = TRUE, spag = FALSE, show.lm = TRUE, show.CI = TRUE, bw = FALSE, B = 1000, quantize = "continuous")
vwReg(outcome ~ predictor, data = df, shade = TRUE, spag = FALSE, show.lm = TRUE, show.CI = TRUE, bw = FALSE, B = 1000, quantize = "SD")
```

### Spaghetti

```{r, message = FALSE, results = "hide"}
vwReg(outcome ~ predictor, data = df, shade = FALSE, spag = TRUE, show.lm = TRUE, show.CI = TRUE, bw = FALSE, B = 1000)
vwReg(outcome ~ predictor, data = df, shade = FALSE, spag = TRUE, show.lm = FALSE, show.CI = FALSE, bw = FALSE, B = 1000)
```

### Black/white

```{r, message = FALSE, results = "hide"}
vwReg(outcome ~ predictor, data = df, shade = TRUE, spag = FALSE, show.lm = TRUE, show.CI = TRUE, bw = TRUE, B = 1000, quantize = "continuous")
vwReg(outcome ~ predictor, data = df, shade = TRUE, spag = FALSE, show.lm = TRUE, show.CI = TRUE, bw = TRUE, B = 1000, quantize = "SD")
vwReg(outcome ~ predictor, data = df, shade = FALSE, spag = TRUE, show.lm = TRUE, show.CI = TRUE, bw = TRUE, B = 1000, quantize = "SD")
```

# Graphic Design Principles for Data Visualization

https://www.data-to-viz.com/caveats.html

# Types of Plots

## Univariate Distribution

Used for: distribution of one numeric variable

### Gallery

- Violin chart: https://r-graph-gallery.com/violin.html
- Density chart: https://r-graph-gallery.com/density-plot.html
- Histogram: https://r-graph-gallery.com/histogram.html
- Boxplot: https://r-graph-gallery.com/boxplot.html
- Ridgeline chart: https://r-graph-gallery.com/ridgeline-plot.html

## Bivariate Scatterplots

Used for: association between two numeric variables

### Base `R`

```
plot(x, y)
```

### ggplot2 package

http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/

```
ggplot(data, aes(x, y)) +
    geom_point()
```

### Gallery

- Scatterplot: https://r-graph-gallery.com/scatterplot.html
- Bubble plot: https://r-graph-gallery.com/bubble-chart.html
- 2D density chart: https://r-graph-gallery.com/2d-density-chart.html
- Heatmap: https://r-graph-gallery.com/heatmap.html

### Add lines

- Line chart: https://r-graph-gallery.com/line-plot.html
- Connected scatterplot: https://r-graph-gallery.com/connected-scatterplot.html
- Visually-weighted regression: https://www.nicebread.de/visually-weighted-watercolor-plots-new-variants-please-vote/
- Use the `vwReg()` function from the `petersenlab` package: https://github.com/DevPsyLab/petersenlab/blob/main/R/vwReg.R

### Area

- Area chart: https://r-graph-gallery.com/area-chart.html
- Stacked area chart: https://r-graph-gallery.com/stacked-area-graph.html
- Streamgraph: https://r-graph-gallery.com/streamgraph.html

## Bivariate Barplots

Used for: association between one categorical variable and one numeric variable (or for depicting the frequency of categories of a categorical variable)

### Gallery

- Barplot: https://r-graph-gallery.com/barplot.html
- Lollipop plot: https://r-graph-gallery.com/lollipop-plot.html

## Multivariate Correlation Matrices

Used for: association between multiple numeric variables

For correlation matrices, I do the following:

1. I use the lab's `cor.table()` function (with `type = "manuscript"`) from the `petersenlab` package to create a correlation matrix.
1. I save the correlation matrix to a `.csv` file.
    a. For example: https://research-git.uiowa.edu/PetersenLab/SRS/SRS-SelfRegulationDevelopment/-/blob/master/Analyses/selfRegulationDevelopment.Rmd#self-regulation-measures
1. I open the .csv file in Excel and create the table in Excel that can be copied and pasted to Word/Powerpoint/etc.

### Correlograms

- `corrplot` package: https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
- `corrgram` package: https://cran.r-project.org/web/packages/corrgram/vignettes/corrgram_examples.html

#### Gallery

https://r-graph-gallery.com/correlogram.html

### Pairs panels

`psych` package: https://personality-project.org/r/psych/help/pairs.panels.html

I depict examples of correlograms and pairs panels here: https://isaactpetersen.github.io/Principles-Psychological-Assessment/factor-analysis-PCA.html#correlations-factorAnalysis

## Path Diagrams {#pathDiagrams}

Used for: SEM/CFA/path analysis

If you are just trying to visualize the results of a SEM model fitted using the `lavaan` package, I recommend the `semPlot` package (http://sachaepskamp.com/semPlot/examples) in `R`.
You can see examples of my implementation here: https://isaactpetersen.github.io/Principles-Psychological-Assessment/sem.html#semModelPathDiagram-sem

If you are trying to create a figure for a paper or poster, you might want something that you can draw and customize yourself.
I use Adobe Illustrator for hand-drawn figures.

You can look at various options below:

- `semPlot` package
- `Adobe Illustrator`
- https://www.diagrams.net
- `Onyx`: https://onyx-sem.com
- `yworks`: https://www.yworks.com
- `Microsoft Visio`: https://www.microsoft.com/en-us/microsoft-365/visio/flowchart-software
- `Microsoft Powerpoint`
- `AMOS`
- `Warppls`
- `Graphviz`: https://graphviz.org
    - has an `R` port—this is what we use for our study flowchart via the `DiagrammeR` package: https://rich-iannone.github.io/DiagrammeR/index.html
- https://app.diagrams.net
- https://github.com/jgraph/drawio-desktop/releases

## Interactive

Gallery: https://r-graph-gallery.com/interactive-charts.html

## Animation

Gallery: https://r-graph-gallery.com/animation.html

## 3D

Gallery: https://r-graph-gallery.com/3d.html

# Color Palettes {#colorPalettes}

- https://colorbrewer2.org
- https://sites.stat.columbia.edu/tzheng/files/Rcolor.pdf (archived at https://perma.cc/8SYT-UC5U)

## Sequential

- https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
- viridis, cividis, etc.: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

## Diverging

- https://colorbrewer2.org/#type=diverging&scheme=BrBG&n=3

## Qualitative/Categorical

- https://colorbrewer2.org/#type=qualitative&scheme=Accent&n=3
- `Polychrome` package: https://stackoverflow.com/a/62939405 (archived at https://perma.cc/3HWM-MMFS)
- `pals` package: https://stackoverflow.com/a/41230685 (archived at https://perma.cc/WH56-HMVD)

Color palettes for color-blindness:

- `Safe` palette from the `rcartocolor` package: https://stackoverflow.com/a/56066712 (archived at https://perma.cc/WUH5-F4Z7)
- Okabe Ito scale: https://stackoverflow.com/a/56066712 (archived at https://perma.cc/WUH5-F4Z7)

### 8 Colors

```{r}
# From here: https://github.com/clauswilke/colorblindr/blob/master/R/palettes.R
# Two color palettes taken from the article ["Color Universal Design" by Okabe and Ito](https://web.archive.org/web/20210108233739/http://jfly.iam.u-tokyo.ac.jp/color/)
# The variant `palette_OkabeIto` contains a gray color, while `palette_OkabeIto_black` contains black instead

palette_OkabeIto <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

pie(rep(1, 8), col = palette_OkabeIto)

palette_OkabeIto_black <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")

pie(rep(1, 8), col = palette_OkabeIto_black)
```

### 12 Colors

`Safe` palette from the `rcartocolor` package: https://stackoverflow.com/a/56066712 (archived at https://perma.cc/WUH5-F4Z7)

```{r}
#from: scales::show_col(carto_pal(12, "Safe"))

c12 <- c(
  "#88CCEE",
  "#CC6677",
  "#DDCC77",
  "#117733",
  "#332288",
  "#AA4499",
  "#44AA99",
  "#999933",
  "#882255",
  "#661100",
  "#6699CC",
  "#888888"
)

pie(rep(1, 12), col = c12)
```

### 25 Colors

https://stackoverflow.com/a/9568659 (archived at https://perma.cc/5ALZ-3AQD)

```{r}
c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)

pie(rep(1, 25), col = c25)
```

### 36 Colors

```{r}
# from: Polychrome::palette36.colors(36)

c36 <- c("#5A5156","#E4E1E3","#F6222E","#FE00FA","#16FF32","#3283FE","#FEAF16","#B00068","#1CFFCE","#90AD1C","#2ED9FF","#DEA0FD","#AA0DFE","#F8A19F","#325A9B","#C4451C","#1C8356","#85660D","#B10DA1","#FBE426","#1CBE4F","#FA0087",
"#FC1CBF","#F7E1A0","#C075A6","#782AB6","#AAF400","#BDCDFF","#822E1C","#B5EFB5","#7ED7D1","#1C7F93","#D85FF7","#683B79","#66B0FF","#3B00FB")

pie(rep(1, 36), col = c36)
```

## Maps

- https://github.com/thomasp85/scico

# Session Info

```{r}
sessionInfo()
```

[^1]: Ask me to give you access.
