---
title: "Hierarchical Linear Modeling"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      comment = "",
                      class.source = "fold-show")
```

To do:

- Add data file and examples

# Preamble

## Install Libraries

```{r, class.source = "fold-hide"}
#install.packages("remotes")
#remotes::install_git("https://research-git.uiowa.edu/PetersenLab/petersenlab.git")
```

## Load Libraries

```{r, message = FALSE, warning = FALSE, class.source = "fold-hide"}
library("lme4")
library("nlme")
```

# Terms

These models go by a variety of different terms:

- hierarchical linear model (HLM)
- multilevel model (MLM)
- mixed effects model
- mixed model

# Overview

https://isaactpetersen.github.io/Principles-Psychological-Assessment/reliability.html#mixedModels

# Estimator: ML or REML

For small sample sizes, restricted maximum likelihood (REML) is preferred over maximum likelihood (ML).
ML preferred when there is a small number (< 4) of fixed effects; REML is preferred when there are more (> 4) fixed effects.
The greater the number of fixed effects, the greater the difference between REML and ML estimates.
Likelihood ratio (LR) tests for REML require exactly the same fixed effects specification in both models.
So, to compare models with different fixed effects with an LR test (to determine whether to include a particular fixed effect), ML must be used.
In contrast to the maximum likelihood estimation, REML can produce unbiased estimates of variance and covariance parameters, variance estimates are larger in REML than ML.
To compare whether an effect should be fixed or random, use REML.
To simultaneously compare fixed and random effects, use ML.

# Linear Mixed Models {#linear}

## `lme4`

```{r}
lmer(t_ext ~ m_ext + age + (age | tcid),
     data = mydata,
     REML = FALSE, #for ML
     na.action = na.exclude)
```

## `nlme`

```{r}
lme(t_ext ~ m_ext + age,
    random= ~1 + age|tcid,
    data = mydata,
    method = "ML",
    na.action = na.exclude)
```

# Generalized Linear Mixed Models {#generalized}

http://glmm.wikidot.com/faq

## `lmer`

```{r}
glmer(t_ext ~ m_ext + age + (age | tcid),
      family = poisson(link = "log"),
      data = mydata,
      na.action = na.exclude)
```

## `MASS`

```{r}
glmmPQL(t_ext ~ m_ext + age,
        random = ~1 + age|tcid,
        family = poisson(link = "log"),
        data = mydata)
```

## `MCMCglmm`

```{r}
MCMCglmm(t_ext ~ m_ext + age,
         random = ~ us(age):tcid,
         family = "poisson",
         data = na.omit(mydata[,c("tcid","t_ext","m_ext","age")]))
```

# Nonlinear Mixed Models {#nonlinear}

```{r}
nlme(height ~ SSasymp(age, Asym, R0, lrc),
     data = Loblolly,
     fixed = Asym + R0 + lrc ~ 1,
     random = Asym ~ 1,
     start = c(Asym = 103, R0 = -8.5, lrc = -3.3))
```

# Robust Mixed Models

To evaluate the extent to which a finding could driven by outliers, this could be done in a number of different ways, such as:

- identifying and excluding influential observations based on DFBETAS or Cook’s distance (Nieuwenhuis, Grotenhuis, & Pelzer, 2012)
- fitting mixed models using rank-based estimation (Bilgic & Susmann, 2013; Finch, 2017) or robust estimating equations (Koller, 2016)
- estimating robust standard errors using a sandwich estimator (Wang & Merkle, 2018)

# Assumptions

The within-group errors:

1. are independent
2. are identically normally distributed
3. have mean zero and variance sigma-squared
4. are independent of the random effects

The random effects:

5. are normally distributed
6. have mean zero and covariance matrix Psi (not depending on the group)
7. are independent for different groups

# Examining Model Assumptions

## Resources

Pinheiro and Bates (2000) book (p. 174, section 4.3.1)

https://stats.stackexchange.com/questions/77891/checking-assumptions-lmer-lme-mixed-models-in-r

## QQ Plots

```{r}
qqnorm(multilevelRegressionModel, ~ranef(., level = 1)) #Multilevel model: make QQ plots for each level of the random effects. Vary the level from 0, 1, to 2 so that you can check the between- and within-subject residuals
```

## PP Plots

```{r}
ppPlot(multilevelRegressionModel)
```

## QQ Plot of residuals

```{r}
qqnorm(resid(multilevelRegressionModel))
```

## Plot residuals

```{r}
plot(multilevelRegressionModel)
```

## Plot residuals by group (in the example below, level 2 represents the individual)

```{r}
plot(multilevelRegressionModel, as.factor(tcid) ~ resid(.), abline = 0)
```

## Plot residuals by levels of a predictor

```{r}
plot(multilevelRegressionModel, resid(., type = "p") ~ fitted(.) | female) #type="p" specifies standardized residuals
```

## Can model heteroscedasticity of the within-group error with the weights argument

```{r}
multilevelRegressionModelVarStructure <- lme(t_ext ~ m_ext + age + female,
                                             random = ~1 + age|tcid,
                                             weights = varIdent(form = ~ 1 | female),
                                             method = "ML",
                                             data = mydata,
                                             na.action = na.exclude)
```

## Plot observed and fitted values

```{r}
plot(multilevelRegressionModelVarStructure, t_ext ~ fitted(.))
```

## Plot QQ plot of residuals by levels of a predictor

```{r}
qqnorm(multilevelRegressionModelVarStructure, ~ resid(.) | female)
qqnorm(multilevelRegressionModelVarStructure, ~ resid(.))
```

## QQ plot of random effects

```{r}
qqnorm(multilevelRegressionModelVarStructure, ~ranef(., level = 1)) #make QQ plots for each level of the random effects. Vary the level from 0, 1, to 2 so that you can check the between- and within-subject residuals
qqnorm(multilevelRegressionModelVarStructure, ~ranef(., level = 0))
qqnorm(multilevelRegressionModelVarStructure, ~ranef(., level = 2))
```

## QQ plot of random effects by levels of a predictor

```{r}
qqnorm(multilevelRegressionModelVarStructure, ~ranef(., level = 1) | female)
```

## Pairs plot

```{r}
pairs(multilevelRegressionModelVarStructure)
pairs(multilevelRegressionModelVarStructure, ~ranef(., level = 1) | female)
```

## Variance functions for modeling heteroscedasticity

- `varFixed`: fixed variance
- `varIdent`: different variances per stratum
- `varPower`: power of covariate
- `varExp`: exponential of covariate
- `varConstPower`: constant plus power of covariate
- `varComb`: combination of variance functions

## Correlation structures for modeling dependence

- `corCompSymm`: compound symmetry
- `corSymm`: general
- `corAR1`: autoregressive of order 1
- `corCAR1`: continuous-time AR(1)
- `corARMA`: autoregressive-moving average
- `corExp`: exponential
- `corGaus`: Gaussian
- `corLin`: linear
- `corRatio`: rational quadratic
- `corSpher`: spherical

# Session Info

```{r, class.source = "fold-hide"}
sessionInfo()
```

