---
title: "High-Performance Computing (Argon Supercomputer)"
---

# Overview

Analysis scripts for High Performance Computing (HPC) at the University of Iowa.
Currently uses Argon and IDAS.

## Argon vs. IDAS

Both Argon and IDAS rely on HPC to run intensive, large-scale scripts and processing jobs.
See the [Argon vs. IDAS comparison page](https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/540606569/Comparing+IDAS+and+Argon) (archived at <https://perma.cc/Q7G5-4JA3>) for more information about which option to use.

Argon runs on CentOS Linux, which requires users to have a basic understanding of command-line interface (CLI).
IDAS is similar to a remote desktop in that users can open and interact with Rstudio, Jupyter Notebook, and Visual Studio Code while benefiting from the higher processing capabilities of the HPC system.
This documentation focuses primarily on Argon.

## Links to Documentation

Workflow to request Argon/HPC access: <https://workflow.uiowa.edu/entry/new/3282>

Workflow to request IDAS access: <https://workflow.uiowa.edu/entry/new/7941>

Cluster Systems Documentation:
<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513411/Cluster+Systems+Documentation> (archived at <https://perma.cc/EKB8-ZKR7>)

Argon Cluster Documentation:
<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513466/Argon+Cluster> (archived at <https://perma.cc/6HST-VV6Y>)

IDAS Documentation: <https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513404/Interactive+Data+Analytics+Service+Documentation> (archived at <https://perma.cc/2HT8-7FWU>)

Beginner Overview of HPC: <https://carpentries-incubator.github.io/hpc-intro/index.html> (archived at <https://perma.cc/Y8GR-QU3W>)

Beginner Overview of Linux Command Line: <https://ubuntu.com/desktop/docs/en/latest/tutorial/the-linux-command-line-for-beginners/> (archived at <https://perma.cc/73G9-B8QK>)

## Data Storage

Significant amounts of data storage are provided on Argon, but data are not backed up in any way unless special arrangements are made.
Standard (free) Argon accounts include 1 TB of storage, and standard IDAS accounts include 100 GB of storage.
It is the responsibility of the user to back up important information.

## Sensitive Data

User agrees to refrain from storing Restricted data on HPC resources.
Data is classified as Restricted when the unauthorized disclosure, alteration or destruction of that data could cause a significant level of risk to the University or its affiliates.
Examples of Restricted data include data protected by state or federal privacy regulations and data pertaining to identified human subjects that has not been deidentified.

## Fees

At present there are no fees for the use of the Argon cluster or IDAS for low-priority usage.
For large users, or those who want access to dedicated resources, the option of purchasing or renting supplemental system hardware may be available.
If you are interested in dedicated hardware, contact [research-computing@uiowa.edu](mailto:research-computing@uiowa.edu).

# Accessing Argon via SSH

SSH, or **S**ecure **SH**ell, is a tool that allows users to access and operate remote servers (i.e., other computers).
You do not need to download or install anything to use SSH.
Users can implement SSH via the terminal or via an emulator software, such as SecureCRT or PuTTY.
Emulators allow users to enter commands into the terminal and often include GUIs that show the file structure of the remote server.
This can be helpful for users who have not used Linux, which operates via command line input and does not have an equivalent to Windows File Explorer or Mac Finder.
With SSH access, you can view and edit files as well as run scripts and submit jobs.

## Using the Terminal

1. Open Command Prompt (Windows) or Terminal (MacOS)
1. If you are on campus and/or using the VPN, enter `ssh hawkid@argon.hpc.uiowa.edu`
   If you are off campus and not using the VPN, enter `ssh -p 40 hawkid@argon.hpc.uiowa.edu`
1. Enter your HawkID password when prompted
    - Note: The individual characters of your password will not appear as you type--this is not a bug!
1. Complete the two-factor authentication

## Using SecureCRT

SecureCRT is the University's choice of emulator, and it has several customizable settings for more advanced users.

1. Download [SecureCRT](https://its.uiowa.edu/securecrt) from UIowa Informational Technology Services: <https://its.uiowa.edu/securecrt>
1. Enter `argon.hpc.uiowa.edu` as the Hostname and your HawkID as the Username
    - Do not modify any other fields/settings
1. Click Connect
1. Enter your HawkID password when prompted
1. Complete the two-factor authentication

## Using PuTTY

PuTTY is another popular emulator that is more lightweight but less customizable than SecureCRT.
It is not available through ITS, but it is free to download from the creators.

1. Download [PuTTY](https://www.chiark.greenend.org.uk/~sgtatham/putty/)
1. Enter `argon.hpc.uiowa.edu` as the Host Name
1. Click Connect
    - The first time you connect, click Accept on the pop-up alert
1. Enter your HawkID after the `login as:` prompt
1. Enter your HawkID password when prompted
1. Complete the two-factor authentication

## After Log-In

After logging in, the terminal or emulator will take a moment to connect to Argon.
Once a connection is established, you will see `(base) [hawkid@argon-itf-login-# ~]$` as the final line of text.

# Map Argon Drive to Local Computer

Mapping your Argon directory to your computer allows you to access your Argon files without CLI.
You won't be able to submit jobs, but it makes editing and moving files much more convenient for those who are new to Linux.

## Windows

1. Open File Explorer
1. In the sidebar on the left, right click 'This PC', then select 'Map Network Drive'
1. Enter `\\data.hpc.uiowa.edu\argon_home` as the Folder
1. Ensure that the 'Reconnect at sign-in' box is checked
1. Click 'Finish'

## Mac

1. Open Finder
1. In the menus along the top, select 'Go' then 'Connect to Server...'
1. Enter `smb://IOWA;<hawkid>:*@data.hpc.uiowa.edu/argon_home` as the Server Address
1. Click 'Connect'

# Argon Navigation

## Quick Reference

- `ssh hawkid@argon.hpc.uiowa.edu` logs in to Argon
- `exit` closes out of whichever node/environment/connection you are in
- `qlogin` requests a computing node
- `pwd` shows the path to the current folder
- `cd insert/file/path` brings you to the provided file path
    - `cd ~` returns to home directory (your Argon folder)
    - `cd ..` returns to the folder "above" the current folder (i.e., the folder that the current folder is in)
    - `cd /Shared/lss_itpetersen/Lab` goes to the lab drive
- `ls` lists the files and folders in your current folder
- `man insertCommandName` opens instructions for the provided command
- `sudo insertCommandName` runs the provided command with administrative privileges
    - If you are not logged in as an administrator, you may be prompted for an admin login

## Nodes

After logging in, you are assigned to a *log-in node*.
There are limited log-in nodes, and they are shared by all Argon users.
You can create, edit, and delete files, and you can submit jobs for processing, but you should not do any actual computing from the log-in node.
To do more intensive work, you will need to request a *computing node* by entering `qlogin` in the terminal or emulator.

You can think of the login nodes as the host stand at a restaurant--you go there when you first arrive (log in) and ask for a table (computing node), and the host (queue manager) seats you when one is available.
However, it would be rude and inconvenient to order your food and eat it (run a bunch of scripts) at the host stand.

## File Systems

To interact with files in Argon, you must be able to navigate through the file directory.
On Windows and MacOS, File Explorer and Finder allow users to locate the files they need.
In Argon (and other Linux-based systems), users navigate through directories (folders) using the command line.
There are a few commands that are helpful for moving through folders:

- `pwd` shows the full file path to the current folder.
    - When you first log in to Argon, you will be in your home directory, `/Users/hawkid`
    - The home directory can be abbreviated to `~`
- `cd insert/file/path` allows you to move to different folders
    - If you are moving to a folder that is somewhere inside of the current folder, you can use the relative file path
        - Example: To move from the main `Lab` folder in the lab drive to the Studies folder, enter `cd Studies`
        - Example: To move from the main `Lab` folder in the lab drive to your Members folder, enter `cd Members/hawkid`
    - If you are moving to a folder that is outside of the current folder, you must use the absolute (full) file path
        - Example: To move from your Argon home directory to the Studies folder on the lab drive, enter `cd /Shared/lss_itpetersen/Lab/Studies`
    - If the folder name has any spaces in it, put quotation marks around the file path
        - Example: `cd "Studies/School Readiness Study"`
    - `cd ~` will bring you back to your home directory (i.e., your Argon storage folder)
    - `cd ..` will bring you to the folder "above" the folder that you are currently in
        - Example: If you are in the `School Readiness Study` folder in the lab drive, `cd ..` will bring you to the `Studies` folder
- `ls` lists all of the files and folders in your current folder
    - All folders contain `.` (the current folder) and `..` (the parent folder)
    - Below are some common arguments used with `ls` (arguments can be combined):
        - `ls -A` lists all files and folders, including hidden ones but excluding `.` and `..`
        - `ls -d` lists folders only
        - `ls -l` lists folder contents with permissions, owner, size, and modification date
        - `ls -R` lists contents of current folder and all subfolders
        - `ls /file/path` lists the contents of the specified folder

### Accessing the Lab Drive

After accessing Argon using SSH, enter `cd /Shared/lss_itpetersen/Lab` in the terminal or emulator.

### Project Directories

Like the Lab drive, it is recommended that you store the materials for each project in its own folder:
```bash
cd /Users/itpetersen/Documents/Projects/Bayesian_IRT/
cd /Users/itpetersen/Documents/Projects/EXT_pilot/
cd /Users/itpetersen/Documents/Projects/EXT_ParentReport/
cd /Users/itpetersen/Documents/Projects/Multiple_Imputation/
cd /Users/itpetersen/Documents/Projects/SelfRegulation_IRT/
cd /Users/itpetersen/Documents/Projects/Test/
```

# Using a Container {#sec-container}

Link to ITS documentation: <https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513442/Singularity+Containers>

Link to Apptainer documentation: <https://apptainer.org/documentation/>

## Overview

Because the default Argon workspace only has a limited set of software/apps available (the 'stack'), many users prefer to use a container.
A container essentially operates like its own computer, with its own operating system and any files or programs the creator wants to include.
However, unlike an actual computer, a container is stored as a file that can be shared (and used) across several different computers.

As an example of why this might be helpful, consider a script that runs in `R`.
Two different users might have different operating systems (Windows, MacOS, or Linux), different versions of `R`, different packages installed, or different versions of the required packages installed.
One user might not have `R` installed at all.
With a container, the user does not have to figure out which software and software versions are necessary to run a certain script--all they need is the script and the container.
Containers are also helpful for reproducibility.
Because the container is stable (i.e., the software and operating system do not update unless the creator modifies it), the results of a given script will be the same each time it is run in that container.

The University recommends using Apptainer, formerly known as Singularity, to create and manage containers.

To create a container, you must have admin privileges on the computer you are using, which means you cannot create a container in Argon.
However, you can create a container on your own computer and use it in Argon.
Because none of the lab computers run on Linux, you will need to install a Linux subsystem, which allows users to run Linux from another operating system (such as Windows or MacOS).

## Creating a Container

### Installing Apptainer

1. Install a Linux subsystem with Ubuntu
    - Note: If a Linux subsystem is already installed, open the subsystem as an administrator and proceed to the next step
    <details><summary>Windows</summary>

    1. Open PowerShell as an administrator
    1. Enter `bash wsl --install`
    1. When the installation is complete, restart your computer
    1. Your computer will install Ubuntu (a flavor/type of Linux) during the restart
    1. Open Ubuntu as an administrator
    </details>
    <details><summary>MacOS (Not Recommended)</summary>

    1. Download the latest [Ubuntu server](https://ubuntu.com/download/server/arm)
    1. Download [UTM](https://docs.getutm.app/installation/macos/)
        1. Use the 'Download from GitHub' option--the App Store version is identical, but costs $10
    1. Open UTM and click '+' to create a new virtual machine (VM)
    1. Select 'Virtualize'
    1. Select 'Linux'
    1. Select the amount of RAM and CPU cores to allocate to the VM (or use defaults)
        - To see how much RAM/CPU cores your computer has:
            1. Click the Apple icon in the top left corner of the screen
            1. Select 'About this Mac'
            1. The RAM is listed next to 'Memory'
            1. Click 'More Info...'
            1. Scroll all the way to the bottom, then click 'System Report...'
            1. The number of CPU cores is listed next to 'Total Number of Cores'
        - In general, you should allocate no more than 75% of your total CPU cores and no more than 50% of your total RAM
    1. When prompted for the Boot Image, make sure 'Boot from ISO image' is selected, then click 'Browse' and locate the Ubuntu installation file
    1. Specify the maximum amount of storage space to allocate to the subsystem
        - To see how much storage space your computer has:
            1. Open System Settings
            1. Click 'Storage'
        - In general, you should overestimate the amount of storage, but you cannot exceed the available storage space on your computer
    1. Do not add a Shared Directory
    1. Click 'Save'
    1. Click the Play icon to enter the subsystem
    1. Select the 'Try or Install Ubuntu' option
    1. Complete the installation prompts, using the up and down arrow keys to move between options and the enter key to select the highlighted option
        <details><summary>Installation Prompts</summary>

        - Language: English (or whichever you prefer)
        - Keyboard Layout: English (US)
        - Keyboard Variant: English (US)
        - Installation Type: Ubuntu Server
        - Network Configuration: name varies, but select the option with the 'eth' type
        - Proxy Configuration: (leave blank)
        - Mirror Configuration: (leave blank; wait for 'This mirror location passed tests.' to appear)
        - Guided Storage Configuration: (use default options)
        - Storage Configuration: (review summary and continue)
        - Confirm Destructive Action: Continue
            - Note: This pop-up is alarming, but (assuming you set an appropriate storage limit) nothing will be destroyed
        - Profile Configuration:
            - Your name: hawkid
            - Your server's name: hawkid_linux_sub
            - Pick a username: hawkid
            - Choose a password/Confirm your password: (HawkID password recommended)
        - Upgrade to Ubuntu Pro: Skip for now
        - SSH Configuration: (do not enable)
        - Featured server snaps: (select none)
        </details>
    1. When the installation is complete, select 'Reboot Now'
    1. After rebooting (on the screen with the 'Please remove the installation medium...' prompt), manually quit the VM by clicking the Power icon in the top left corner
    1. Click the Play icon to restart the VM
    1. Select the 'Boot from next volume' option
    1. Log in with the username and password created during installation
    </details>
1. Update Ubuntu
    1. Enter `sudo apt update` to check for updates
    1. Enter `sudo apt upgrade -y` to install the updates identified in the previous step
1. Install Apptainer
    1. Run the following code to install dependencies:
        ```bash
        sudo apt update
        sudo apt install -y \
        software-properties-common \
        squashfuse \
        fuse2fs \
        gocryptfs
        ```
    1. Run the following code to download and install Apptainer:
        ```bash
        sudo add-apt-repository -y ppa:apptainer/ppa
        sudo apt update
        sudo apt install -y apptainer
        ```
        - If Apptainer is not available via apt (`E: Unable to locate package apptainer`), install it from source
            <details><summary>Source Installation</summary>

            1. Get the most current version number from [GitHub](https://github.com/apptainer/apptainer/releases) (do not need to download--just need version number)
            1. Define a variable to hold version info (insert version number in place of #s):
                ```bash
                export VERSION=#.#.#
                ```
            1. Download the source code:
                ```bash
                wget https://github.com/apptainer/apptainer/releases/download/v${VERSION}/apptainer-${VERSION}.tar.gz
                tar -xzf apptainer-${VERSION}.tar.gz
                cd apptainer-${VERSION}
                ```
            1. Build Apptainer:
                ```bash
                ./mconfig
                make -C builddir
                sudo make -C builddir install
                ```
            </details>
    1. Verify installation:
        ```bash
        apptainer --version
        ```
1. Go to your home directory (`cd ~`)
1. Create a `containers` folder if one does not exist:
    ```bash
    mkdir containers
    ```
    
### Definition Files

Definition files contain the necessary information for installing the container software (e.g., versions, packages).
Because the user (i.e., you) specifies the content of the definition file, there is a lot of flexibility when it comes to what to include and how to structure them.
However, there are a handful of necessary components for a definition file, which you can read about [here](https://apptainer.org/docs/user/latest/definition_files.html#definition-files).

There are a number of pre-written definition files for `R` (with or without `brms` and `CmdStan`), `MPlus`, `MATLAB`, and `Python`/`MNE`, which you can view or download [here](https://research-git.uiowa.edu/PetersenLab/HPC/-/tree/master/Argon/Containers/DefinitionFiles).
There are also instructions below for creating a definition file from scratch.
To build a container with a program that is not included here, see the Apptainer documentation for [building a container](https://apptainer.org/docs/user/latest/build_a_container.html) and for [including multiple apps in one container](https://apptainer.org/docs/user/latest/definition_files.html#scif-apps).

1. Go to the `containers` folder (`cd containers`)
1. If installing `MPlus`:
    1. Copy the `MPlus` installer into your `containers` folder:
        ```bash
        cd path/to/installer/MplusLinux64.bin ~/containers/
        ```
    1. Run the installer to generate a properties file:
        ```bash
        ./MplusLinux64.bin -r ~/containers/mplus_install.properties
        ```
    1. Open the properties file:
        ```bash
        nano mplus_install.properties
        ```
    1. Under the "Choose Install Folder" and "Install" headings, update the installation path to `/opt/mplus`
        <details><summary>Example Properties File</summary>
        
        ```bash
        # Tue Feb 24 21:04:04 CST 2026
        # Replay feature output
        # ---------------------
        # This file was built by the Replay feature of InstallAnywhere.
        # It contains variables that were set by Panels, Consoles or Custom Code.
        #Choose Install Set
        #------------------
        CHOSEN_FEATURE_LIST=Application,Examples
        CHOSEN_INSTALL_FEATURE_LIST=Application,Examples
        CHOSEN_INSTALL_SET=Typical
        #Choose Install Folder
        #---------------------
        USER_INSTALL_DIR=/opt/mplus
        #Install
        #-------
        -fileOverwrite_/opt/mplus/uninstall/uninstall_mplus.lax=Yes
        -fileOverwrite_/opt/mplus/mplus=Yes
        -fileOverwrite_/opt/mplus/Documentation/Getting_Started.pdf=Yes
        -fileOverwrite_/opt/mplus/Documentation/Mplus_Users_Guide_v8.pdf=Yes
        -fileOverwrite_/opt/mplus/Documentation/Version_8.1_Language_Addendum.pdf=Yes
        -fileOverwrite_/opt/mplus/Documentation/Version_8.5_Language_Addendum.pdf=Yes
        ```
        </details>
    1. Press `Ctrl+O` then `Enter` to save ("write out") the file
    1. Press `Ctrl+X` to close the file
1. If installing `MATLAB`:
    1. Copy the dependencies file into your `containers` folder:
        ```bash
        wget -q -O https://raw.githubusercontent.com/mathworks-ref-arch/container-images/refs/heads/main/matlab-deps/r2026a/ubuntu24.04/base-dependencies.txt
        ```
1. Create or download the definition file
    1. To download an existing definition file (see [GitLab](https://research-git.uiowa.edu/PetersenLab/HPC/-/tree/master/Argon/Containers/DefinitionFiles) for options, or ):
        1. Enter the following code into the terminal:
            ```bash
            wget -q -O argon_container.def https://link.to.definition.file
            ```
        1. Log in with your HawkID credentials when prompted
    1. To create a definition file from scratch:
        1. Create a new file:
            ```bash
            nano argon_container.def
            ```
        1. Add the definition file header:
            ```bash
            Bootstrap: docker
            From: ubuntu:24.04
            ```
            Other common base images:
              - `anaconda/miniconda:latest` for a container with `miniconda` (helpful for `Python` environments)
              - `mathworks/matlab` for `MATLAB` only
              - `rocker/r-ver:latest` for `R`, `CmdStan`, and/or `MPlus`
                  - Note: The definition file below uses r2u, which automatically installs Linux dependencies for R packages.
                  The base `R` image at `rocker/r-ver` does not include r2u, so you will have to add it in the definition file or manually install the Linux dependencies.
        1. Add remaining parameters.
            The script below includes parameters for `R` (with `brms`), `CmdStan`, `MPlus`, `MATLAB`, and `Python`/`MNE`.
            Paste the entire script into the definition file.
            To install only a subset of the software, simply remove the section(s) associated with the unwanted software.
            Note that `MPlus` has a subsection in the `R` section in addition to its standalone section.
            Also note `CmdStan` does not have a standalone app section--it is installed as a subsection of the `R` installation.
              
            <details><summary>All Software</summary>
            
              ```bash
              %environment
                
                  # MPlus paths
                  export PATH=/opt/mplus:$PATH
                  
                  # R paths
                  export R_LIBS_USER=/usr/local/lib/R/site-library
                  
                  # CmdStan paths
                  export CMDSTAN=/opt/cmdstan
                  export STAN_THREADS=true
                  
              %files
                  
                  # MPlus files
                  MplusLinux64.bin /opt/MplusLinux64.bin
                  mplus_install.properties /opt/mplus_install.properties
                  
                  # MATLAB files
                  base-dependencies.txt /opt/base-dependencies.txt
              
              %post
              
                  # install dependencies for all apps (do not delete)
                  # includes any dependencies required for 2+ software packages
                  apt-get update && apt-get install -y --no-install-recommends \
                    bzip2 \
                    ca-certificates \
                    cmake \
                    curl \
                    g++ \
                    git \
                    make \
                    nano \
                    python3 \
                    vim \
                    wget

                  # clean up package manager
                  apt-get clean -y
                  apt-get autoremove -y
                  rm -rf /var/lib/apt/lists/*
                  
              #############################################
              # Python/MNE
              #############################################
              
              %apprun python
              
                  exec python "$@"

              %appinstall python
              
                  # install pip
                  apt-get update && apt-get install -y --no-install-recommends \
                    python3-pip
                  
                  # install mne, overriding env warning
                  pip install --break-system-packages "mne[full]"
                  
                  # install hdf5 (large file handling) support packages, overriding env warning
                  pip install --break-system-packages h5io pymatreader
              
              #############################################
              # R
              #############################################
              
              %apprun R
              
                  exec R "$@"
              
              %appinstall R
              
                  # install additional r2u dependencies
                  # if installing R without r2u (from rocker base), add Linux deps here
                  # gnupg required for r2u, all others only needed for eegUtils package
                  apt-get update && apt-get install -y --no-install-recommends \
                    gnupg \
                    liblapack-dev \
                    libblas-dev \
                    gfortran
                  
                  # add r2u keys to apt folder so they are recognized as safe
                  gpg --homedir /tmp --no-default-keyring \
                    --keyring /usr/share/keyrings/r2u.gpg \
                    --keyserver keyserver.ubuntu.com \
                    --recv-keys A1489FE2AB99A21A 67C2D66C4B1D4339 51716619E084DAB9
                  
                  # add r2u mirror to apt source list to pull files
                  # do not change indentation - I know it's ugly but it will throw an error
                  cat > /etc/apt/sources.list.d/r2u.sources <<EOF
              Types: deb
              URIs: https://r2u.stat.illinois.edu/ubuntu
              Suites: noble
              Components: main
              Arch: amd64, arm64
              Signed-By: /usr/share/keyrings/r2u.gpg
              EOF
              
                  # add base r to apt source list just to be safe
                  cat > /etc/apt/sources.list.d/cran.sources <<EOF
              Types: deb
              URIs: https://cloud.r-project.org/bin/linux/ubuntu
              Suites: noble-cran40/
              Components:
              Arch: amd64, arm64
              Signed-By: /usr/share/keyrings/r2u.gpg
              EOF
                  
                  # install base r
                  apt-get update && apt-get install -y --no-install-recommends r-base-core
                  
                  # pin r2u to top of package priority list so it gets picked for installs
                  # don't change indentation here either
                  cat > /etc/apt/preferences.d/99cranapt <<EOF
              Package: *
              Pin: release o=CRAN-Apt Project
              Pin: release l=CRAN-Apt Packages
              Pin-Priority: 700
              EOF
                  
                  # install python dependencies for bspm
                  apt-get update && apt-get install -y --no-install-recommends \
                    python3-dbus \
                    python3-gi \
                    python3-apt
                    
                  # install bspm
                  sudo Rscript --vanilla -e "install.packages(
                    'bspm',
                    repos = 'https://cloud.r-project.org'
                  )"
                  
                  # update bspm settings
                  # don't change indentation
                  mkdir -p /etc/R
                  cat >> /etc/R/Rprofile.site <<EOF
              suppressMessages(bspm::enable())
              options(bspm.version.check = FALSE)
              EOF
                  
                  # install packages (add/remove as needed)
                  R -e "install.packages(
                    c(
                      'brms',
                      'data.table',
                      'devtools',
                      'flowchart',
                      'fs',
                      'ggh4x',
                      'gt',
                      'gtsummary',
                      'janitor',
                      'lavaan',
                      'lme4',
                      'parallely',
                      'performance',
                      'psych',
                      'R.matlab',
                      'remotes',
                      'renv',
                      'semTools',
                      'tidyverse'
                    ),
                    repos = 'https://cloud.r-project.org'
                  )"
                  
                  # install packages via remotes
                  R -e "remotes::install_github('DevPsyLab/petersenlab')"
                  R -e "remotes::install_github('craddm/eegUtils')"
                  
                  # clean up package manager
                  apt-get clean -y
                  apt-get autoremove -y
                  rm -rf /var/lib/apt/lists/*
                  
                  #######
                  # Mplus
                  #######
                  R -e "install.packages(
                    'MplusAutomation',
                    repos = 'https://cloud.r-project.org'
                  )"
                  
                  #############
                  # CmdStan
                  #############
                  
                  # install R package
                  R -e "install.packages(
                    'cmdstanr',
                    repos = c(
                      'https://stan-dev.r-universe.dev',
                      'https://cloud.r-project.org'
                    )
                  )"
                  
                  # create folder to install into
                  mkdir -p /opt/cmdstan
                  
                  # install CmdStan using R
                  R -e "cmdstanr::install_cmdstan(
                    cores = 2,
                    dir = '/opt/cmdstan'
                  )"
                  
                  # create an R environment file
                  mkdir -p /root/.R
                  
                  # add threading vars to environment and startup
                  echo 'STAN_THREADS=true' >> /root/.Renviron
                  echo 'STAN_THREADS=true' >> /root/.R/Makevars

                  # clean up package manager
                  apt-get clean -y
                  apt-get autoremove -y
                  rm -rf /var/lib/apt/lists/*
                  
              #############################################
              # MPlus
              #############################################
              
              %apprun mplus
              
                  exec mplus "$@"
                  
              %appinstall mplus
              
                  chmod +x /opt/MplusLinux64.bin
                  /opt/MplusLinux64.bin -i silent -f /opt/mplus_install.properties
                  chmod -R 755 /opt/mplus
                  
              #############################################
              # MATLAB
              #############################################
              
              %apprun matlab
              
                  exec matlab -nodesktop -nojvm -nosplash "$@"
                  
              %appinstall matlab
              
                  # define var for noninteractive install
                  DEBIAN_FRONTEND=noninteractive \
                  
                  # install additional dependencies
                  apt-get update && apt-get install -y --no-install-recommends \
                    `cat /opt/base-dependencies.txt`
                    
                  # clean up package manager
                  apt-get clean -y
                  apt-get autoremove -y
                  rm -rf /var/lib/apt/lists/*
              
                  # define variables needed for setup
                  export MATLAB_RELEASE=R2026a
                  export MATLAB_PRODUCT_LIST="MATLAB Wavelet_Toolbox Signal_Processing_Toolbox Statistics_and_Machine_Learning_Toolbox Optimization_Toolbox Parallel_Computing_Toolbox"
                  export MATLAB_INSTALL_LOCATION="/opt/matlab/${MATLAB_RELEASE}
                  
                  # add those variables to apptainer environment
                  # do not change the line indentation - it will cause an error
                  cat << EOF > $APPTAINER_ENVIRONMENT
              export MATLAB_RELEASE=$MATLAB_RELEASE
              export MATLAB_PRODUCT_LIST=$MATLAB_PRODUCT_LIST
              export MATLAB_INSTALL_LOCATION=$MATLAB_INSTALL_LOCATION
              EOF
                  
                  # make sure apptainer has full permissions in tmp
                  chmod 777 /tmp
                  
                  # get matlab package manager (mpm) and update permissions
                  wget -q https://www.mathworks.com/mpm/glnxa64/mpm
                  chmod +x mpm
                  
                  # run mpm to install matlab
                  ./mpm install \
                    --release=${MATLAB_RELEASE} \
                    --destination=${MATLAB_INSTALL_LOCATION} \
                    --products ${MATLAB_PRODUCT_LIST}
                  
                  # clean up after installer
                  rm -f mpm /tmp/mathworks_root.log
                  ln -s ${MATLAB_INSTALL_LOCATION}/bin/matlab /usr/local/bin/matlab
                  
                  # add license info
                  mkdir -p ${MATLAB_INSTALL_LOCATION}/licenses
                  cat << EOF > ${MATLAB_INSTALL_LOCATION}/licenses/network.lic
              SERVER matlab-lm.iowa.uiowa.edu 27000
              USE_SERVER
              EOF
              ```
            </details>
        1. Press `Ctrl+O` then `Enter` to save the file
        1. Press 'Ctrl+X' to close the file
1. Build the container:
    ```bash
    sudo apptainer build argon_container.sif argon_container.def
    ```
    You may be prompted to enter your (admin login) password.
    It will take a few minutes to build, and a lot of text will appear in the terminal.
1. Test the container.
    If you used a definition file without an app structure (i.e., no `%appinstall` sections in the definition file), use `apptainer exec argon_container.sif path/to/software/exec_file` instead of `apptainer run --app app_name argon_container.sif` to start the program.
    1. `Python`/`MNE`
        1. Start the `Python` app within the container:
            ```bash
            apptainer run --app python argon_container.sif
            ```
            If the app is working, you should see the `Python` start-up message, followed by `>>>` in the final line.
        1. Verify the `MNE` installation:
            ```python
            import mne; mne.sys_info()
            ```
            You should see a printout of the system information, including a list of installed libraries.
        1. Enter `exit()` to leave the container.
    1. `R`
        1. Start the `R` app within the container:
            ```bash
            apptainer run --app R argon_container.sif
            ```
            If the app is working, you should see the `R` start-up message, followed by `>` in the final line.
            Containers with `r2u` may show a warning about D-Bus services--this will not affect functionality.
        1. Check `R` version:
            ```r
            version
            ```
            You should see a printout of the version information.
        1. Check package installations:
            ```r
            installed.packages()
            ```
            Ensure that all desired packages have been installed.
        1. Enter `q()` to leave the container.
    1. `MPlus`
        1. Start the `MPlus` app within the container:
            ```bash
            apptainer run --app mplus argon_container.sif
            ```
    1. `MATLAB`
        1. Start the `MATLAB` app within the container:
            ```bash
            apptainer run --app matlab argon_container.sif
            ```
            If the app is working, you should see the `MATLAB` start-up message, followed by `>` in the final line.
            Because `MATLAB` verifies its license via network connection, you must be connected to UIowa Wi-Fi (or be using the VPN) for it to work.
        1. Check `MATLAB` version:
            ```bash
            ver
            ```
            You should see a printout of the version information, including installed products.
        1. Enter `exit` to leave the container.
1. Copy the container to your computer:
    ```bash
    cp ~/containers/r_argon.sif "/mnt/c/Users/hawkid/path/to/folder/"
    ```
    The `/mnt/c` path is equivalent to your computer's `C:` drive, so the rest of the file path should be defined from there.
1. If your admin login is different from your HawkID, restart the terminal (or emulator) to run it as a regular user.
1. Copy the container to Argon, either by using the terminal or by manually moving it in File Explorer/Finder.
1. Test the container on Argon, using the same steps described above.
1. If you installed `CmdStan`, copy the executable files to a writable location:
    ```bash
    mkdir -p $HOME/cmdstan_cache
    apptainer exec /Users/hawkid/path/to/container/argon_container.sif \
      cp -r /opt/cmdstan/. $HOME/cmdstan_cache
    ```
1. To submit a job using the container, use the `apptainer exec` command:
    ```bash
    # run an R script using a container built with %app sections
    /usr/bin/apptainer exec --app R /Users/hawkid/container_path/argon_container.sif my_script.R
    
    # run a MATLAB script using a container built without %app sections
    /usr/bin/apptainer exec /Users/hawkid/container_path/argon_container.sif \
    matlab -nodesktop -nojvm -nosplash my_script.mat
    ```
    ```
    For an example job script using a container, see @sec-jobScript.

# Linux

## General Commands

- `pwd` print current working directory
- `cd path/to/directory` sets working directory
- `ls` print files in directory
- `module load moduleName` loads modules required for analyses; necessary to complete prior to submitting jobs involving R
- `module list` lists downloaded modules
- `qsub` submit a job script to Argon processing
- `R CMD BATCH` submit a single R script to Argon processing
Argon requires Linux-compatible file endings, which is problematic for files created outside Argon (using DOS or CRLF file endings). Use the following commands to check if your files are Argon compatible (i.e., Unix LF) and resolve incompatible file endings:
- `file myfile.job | grep CRLF` checks if a file ending uses the incompatible CRLF format. Remember to check the ending of your file; this script assumes it is .job
- `dos2unix myfile.job` converts CRLF file endings to Argon-compatible LF endings
- `find ~/DirectoryName -name '*job' -exec dos2unix "{}" \;` converts all files in a directory that end with .job to LF format; change `*job` to change other file types

## Initial run

### Load the Latest Stack

`module load stack`

### Stack 2022.2

```bash
module load stack/2022.2
module load r/4.2.2_gcc-9.5.0
cd path
```

### Installing Linux Packages to Install `R` Packages

```bash
module load stack/2022.2
module load r/4.2.2_gcc-9.5.0
module load geos/3.9.1_gcc-9.5.0
module load gdal/2.4.4_gcc-9.5.0
module load proj/5.2.0_gcc-9.5.0
module load gsl/2.7.1_gcc-9.5.0
module load nlopt/2.7.0_gcc-9.5.0
module load jags/4.3.0_gcc-9.5.0-dev
module load zlib/1.2.13_gcc-9.5.0-dev
module load cmake/3.25.0_gcc-9.5.0
module load glpk/4.65_gcc-9.5.0-dev
module load libxml2/2.10.3_gcc-9.5.0
module load r-nloptr
module load r-zlibbioc/1.44.0_gcc-9.5.0
module load r-data-table/1.14.4_gcc-9.5.0
module load r-stringi/1.7.8_gcc-9.5.0
module load r-selectr/0.4-2_gcc-9.5.0
module load r-generics/0.1.3_gcc-9.5.0
module load r-fansi/1.0.3_gcc-9.5.0
module load r-utf8/1.2.2_gcc-9.5.0
module load r-pkgconfig/2.0.3_gcc-9.5.0
module load r-gtable/0.3.1_gcc-9.5.0
module load r-scales/1.2.1_gcc-9.5.0
module load r-tzdb/0.3.0_gcc-9.5.0
module load r-timechange/0.1.1_gcc-9.5.0
module load r-dbi/1.1.3_gcc-9.5.0

```

## Load Software Stacks

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513440/Argon+Software+List> (archived at <https://perma.cc/WJ4Q-GDUS>)

`module load stack/2022.2`

## Job Script {#sec-jobScript}

### Using A Container

```bash
#!/bin/sh

# Set working directory
cd /Users/itpetersen/Documents/Projects/Bayesian_IRT/

# Specify qsub options
#$ -pe smp 4
#$ -M isaac-t-petersen@uiowa.edu
#$ -m eas
#$ -l mf=512G
#$ -l h_vmem=512G
#$ -cwd
#$ -q UI-HM
#$ -e /Users/itpetersen/Documents/Projects/Bayesian_IRT/Output/
#$ -o /Users/itpetersen/Documents/Projects/Bayesian_IRT/Output/

# Load the environment modules
module purge

# Run the R script inside container
apptainer exec \
    --bind $HOME/cmdstan_cache:/opt/cmdstan \
    --cleanenv \
    /Users/itpetersen/Documents/Containers/r_argon.sif \
    Rscript ./Analyses/factorScores.R
```

### Using Software Stacks

```bash
#!/bin/sh

# Set working directory
cd /Users/itpetersen/Documents/Projects/Bayesian_IRT/

# Specify qsub options
#$ -pe smp 4
#$ -M isaac-t-petersen@uiowa.edu
#$ -m eas
#$ -l mf=512G
#$ -l h_vmem=512G
#$ -cwd
#$ -q UI-HM
#$ -e /Users/itpetersen/Documents/Projects/Bayesian_IRT/Output/
#$ -o /Users/itpetersen/Documents/Projects/Bayesian_IRT/Output/

# Load the environment modules
module purge
module load stack/2022.2
module load r/4.2.2_gcc-9.5.0

# Run the R script
Rscript ./Analyses/factorScores.R
```

### qsub options {#sec-qsubOptions}

`-pe smp 4`: specify a parallel environment and number of slots (CPU cores) to be used (`smp` = shared memory parallel environment); increasing the number of slots increases the memory available for the job.
If your job gets killed due to insufficient memory, you should increase the number of slots requested (and, as needed, the `mf` and `h_vmem` flags, below).
You can estimate the memory available per slot by dividing a node's total memory by its total core count.
You can find the specific configurations for each node type on the wiki: <https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513468/Queues+and+Policies>
For instance, if you are using the UI queue that has 512GB of memory spread across 128 cores, each core comes with 4GB of memory (i.e., $512 / 128 = 4$).
Thus, if your job needs 20GB of memory, you would need to request at least 5 cores (i.e., $20 / 4 = 5$).

- The `OMP_NUM_THREADS` variable is set to '1' by default.
If your code can take advantage of the threading then specify `OMP_NUM_THREADS` to be equal to the number of job cores per node requested.

`-M isaac-t-petersen@uiowa.edu`: Set the email address to receive email about jobs.
This must be your University of Iowa email address.

`-m eas`: Specify when to send an email message (; ; ; ; )

- `b` = beginning of job
- `e` = end of job
- `a` = when job is aborted
- `s` = when job is suspended
- `n` = no mail is sent

`-l mf=512G`: request a particular quantity of memory you expect to use (to be available for your computation to start; the request is only applicable at scheduling time.)
The `mem_free` flag tells the scheduler to find a node that has at least that much memory available at the moment your job starts.
However, it does not reserve or lock that memory for your job's exclusive use while it is running.
To ensure your job has a guaranteed memory allocation, you should increase the number of slots requested (see above: i.e., `-pe smp <slots>`).

`-l h_vmem=512G`: request the maximum quantity of virtual memory you expect to need; `h_vmem` is a hard limit.
It acts as a ceiling.
If your job exceeds this amount of memory, the system will terminate the process to protect the node's stability.
However, it does not reserve or lock that memory for your job's exclusive use while it is running.
To ensure your job has a guaranteed memory allocation, you should increase the number of slots requested (see above: i.e., `-pe smp <slots>`).

`-cwd`: Determines whether the job will be executed from the current working directory.
If not specified, the job will be run from your home directory.

`-q UI-HM`: specify queue

`-e /Users/itpetersen/Documents/Projects/Bayesian_IRT/Output/`: Name of a file or directory for standard error.

`-o /Users/itpetersen/Documents/Projects/Bayesian_IRT/Output/`: Name of a file or directory for standard output.

## Submit Job

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513450/Basic+Job+Submission> (archived at <https://perma.cc/2SS2-LEJR>)

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513452/Advanced+Job+Submission> (archived at <https://perma.cc/8H6G-2M2F>)

`cd path/to/dataSet123`

[cd /Users/itpetersen/Documents/Projects/Test/Jobs]

`qsub myscript.job`

Job dependency (run `Job B` when `Job A` is finished):

`qsub -hold_jid JOB_ID test_B.job`

### Bayesian IRT

`cd /Users/itpetersen/Documents/Projects/Bayesian_IRT/Jobs`

`qsub bayesianIRT.job`

`qsub -hold_jid JOB_ID factorScores.job`

### Multiple Imputation

`cd /Users/itpetersen/Documents/Projects/Multiple_Imputation/Jobs`

`qsub srs_selfRegulation.R`

## Queues

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513468/Queues+and+Policies> (archived at <https://perma.cc/UUR7-XLBZ>)

`UI`

`UI-HM`

`UI-GPU-HM`

`UI-DEVELOP`

`all.q`

## Queue Status

`qstat -g c`

`qstat -g c -q QUEUE_NAME`

`qstat -g c -q UI`

## Job Status

`qstat -u itpetersen`

`qstat -j JOB_ID`

## Job Time and Memory

- Terminated jobs: `qacct -j JOB_ID`
- Jobs in progress: `qstat -j JOB_ID | grep usage`

## Cancel Job

`qdel JOB_ID`

## Run `R` Script

See [here](#sec-runRscript)

# R

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76514707/R+Programs+in+Batch+mode+for+HPC> (archived at <https://perma.cc/99JQ-43ZG>)

## Find which versions of `R` are installed

`module spider R`

But, there may be more recent version of `R` installed in the "Additional Software Stacks" (<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513440/Argon+Software+List>; archived at <https://perma.cc/WJ4Q-GDUS>)

## Compile `R`

If you want to compile a more recent version of `R` than is available in the software stacks, see [here](https://brandonlebeau.org/2022/06/02/compile-hpc-iowa/) (archived at <https://perma.cc/C6EX-EZL4>).
However, we now compile the most recent version of `R` (and packages) using a [container](#sec-container).

## Install `R` packages

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76514803/Adding+R+programs+to+a+personal+library> (archived at <https://perma.cc/3SRR-2JE7>)

1. first load the `R` environment module
    - `module load r/4.2.2_gcc-9.5.0`
1. launch R
    - `R`
1. then install the package, (the code below uses a generic `package_name`):
    - `install.packages("package_name", repos = "http://cran.r-project.org", dependencies = TRUE, type = "source", Ncpus = 40)`
    - If using `packrat`, it is preferable to install packages by source, if possible, but you can remove `type = source` if you want to install binaries
    - Assuming that you do not have a personal library directory you will see

    ```r
    Warning in install.packages("package_name", repos = "http://cran.r-project.org") :
    lib = "/opt/R/3.0.2/lib64/R/library"' is not writable

    Would you like to use a personal library instead?  (y/n)
    ```

1. Select `y`
1. Select `y` again when prompted to create the directory; your package should download and install into your personal library directory

Example:

```{r}
#| eval: false

install.packages(c(
    "renv","psych","tidyverse","data.table","nlme","lme4","mirt","TeachingDemos","Amelia","mice","miceadds","abind","future","lavaan","blavaan","Rcpp","igraph","shinystan","StanHeaders","brms","rstan","rjags"),
    repos = "http://cran.r-project.org",
    dependencies = TRUE,
    Ncpus = 40)
```

### Install `R` package locally from source

`install.packages(path_to_file, repos = NULL, type = "source", Ncpus = 40)`

Example:

```{r}
#| eval: false

install.packages(c(
    "renv","psych","tidyverse","data.table","nlme","lme4","mirt","TeachingDemos","Amelia","mice","miceadds","abind","future","lavaan","blavaan","Rcpp","igraph","shinystan","StanHeaders","brms","rstan","rjags"),
    repos = "http://cran.r-project.org",
    type = "source",
    dependencies = TRUE,
    Ncpus = 40)
```

### Managing packages for a project using `renv` {#sec-renv}

1. Create a directory where you want you want to store the project.
1. `cd` to the above directory
1. Start R
1. Load the `renv` package: `library("renv")`
1. At the `R` prompt, initialize the `renv` project on the local repository of `R` packages with:
    - `renv::init(project = "/Users/itpetersen/Documents/Projects/SelfRegulation_IRT/")`
    - or `renv::init()` if you are in the intended working directory
1. You must restart your `R` session in the given project directory after running init in order for the changes to take effect!
1. From this point on you are working in a `renv` project.
Installed packages will go into a library within this project.
After initializing the `renv` project on the local repository of `R` packages, packages from the local repository can be installed with `renv::install()`:
    - `renv::install("package_name")`
1. To restart a `renv` project, simply start `R` from the directory created in step (1).
The project will initialize automatically.

To update version of `renv`: `renv::upgrade()`

To install packages: `renv::install("package_name")`

To update packages: `renv::update()`

To save the current state of your library: `renv::snapshot()`

To restore the state of your library from the lock file: `renv::restore()`

To disable `renv` on a project: `renv::deactivate()`

#### To control which packages to install with `renv` using a `DESCRIPTION` file

If you want to control which packages are installed in a `renv` project, you can use a `DESCRIPTION` file to specify the packages that should be installed:

1. Create a `DESCRIPTION` file in the project directory with the following format:

    ```yaml
    Type: project
    Description: My project.
    Depends:
        packageName1,
        packageName2,
        packageName3
    ```

1. Run `renv::settings$snapshot.type("explicit")` to suppress dependency discover and to enable "explicit" mode: <https://rstudio.github.io/renv/reference/dependencies.html#explicit-dependencies>
1. Run `renv::init(bare=TRUE)` to initialize the project without attempting to discover and install `R` package dependencies

### To install a new package for a project after `renv` has been initialized

1. First, load the `R` environment module (see `Install R packages` section above)
1. cd to the directory of the `R` project
1. launch R
    - `R`
1. then install the package, (the code below uses a generic `package_name`):
    - `install.packages("package_name", repos = "http://cran.r-project.org", type = "source", dependencies = TRUE, Ncpus = 40)`
    - `install.packages(c("renv","psych","tidyverse","data.table","nlme","lme4","mirt","TeachingDemos","Amelia","mice","miceadds","abind","future","lavaan","blavaan","Rcpp","igraph","shinystan","StanHeaders","brms","rstan","rjags","renv"), repos = "http://cran.r-project.org", type = "source", dependencies = TRUE, Ncpus = 40)`
1. `renv::snapshot()`

## Update Packages

```{r}
#| eval: false

update.packages(ask = FALSE)
```

## Run `R` Script {#sec-runRscript}

`module load r/4.0.5_gcc-9.3.0`

`cd path/to/dataSet123`

[cd /Users/itpetersen/Documents/Projects/Test/Analyses]

`Rscript path/to/program.R`

[Rscript test.R]

## Quit R

`q()`

# Stan

<https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Linux> (archived at <https://perma.cc/89H9-L8S6>)

Makevars:

`\\data.hpc.uiowa.edu\argon_home\.R\Makevars`



# Troubleshooting

## Running out of Memory {#sec-outOfMemory}

May need to request more memory, either by requesting more slots (to ensure more memory; i.e., `-pe smp <SLOTS>`), by increasing how much memory is available for the node at the moment your job starts (i.e., `mf=<SIZE>`), and/or by increasing the hard limit for virtual memory (i.e., `h_vmem=<SIZE>`); see @sec-qsubOptions.

To see how much memory was used by a killed job, run `qacct -j JOBID` and look at: `maxvmem`.

## `job [NUMBER] died through signal TERM (15)`

May have run out of memory.
See @sec-outOfMemory.

## `job [NUMBER] died through signal KILL (9)`

May have run out of memory.
See @sec-outOfMemory.

## Exit Status: 143

May have run out of memory.
See @sec-outOfMemory.

## Exit Status: 137

May have run out of memory.
See @sec-outOfMemory.

## Having trouble installing packages?

Sometimes Argon will fail to install packages in an R workspace.
We advise the following steps:

1. Sometimes R provides an error stating something like "fatal error: modulename: No such file or directory".
In this instance, you may wish to exist R with q() and load all the modules that begin with the listed module name, as listed above in "Installing Linux Packages to Install R Packages".
Then reopen R and try to install the package again.
1. You can also try installing the package from binary using the RStudio Package Manager (RSPM):

    ```r
    remotes::install_github("cran4linux/rspm")
    rspm::enable()
    install.packages("PACKAGE_NAME")
    ```

1. If that does not work, you can try downloading the .tar file directly from the CRAN repository. Then copy the file into your Argon folder and type 'install.packages("/Users/path/to/directory/package_name", repos = NULL, type = "source")'.
Note that you may need to download an older version of the package (e.g., from the [CRAN Archive](https://cran.r-project.org/src/contrib/Archive/)), such that it is compatible with the version of R you are running on Argon (which it typically not the most recent R version):

    ```r
    install.packages("http://cran.r-project.org/src/contrib/Archive/MASS/MASS_7.3-60.0.1.tar.gz", repos = NULL, type = "source")
    ```

1. Check if you are installing the necessary modules within your job script: 

    ```r
    module load stack/2022.2
    module load r/4.2.2_gcc-9.5.0
    ```
1. Add a line to your R script to make sure it is accessing the same repository where you installed the packages manually:

    ```r
    .libPaths(c("/old_Users/HAWKID/Rlibs", 
            "/cvmfs/argon.hpc.uiowa.edu/2022.2/apps/linux-centos7-broadwell/gcc-9.5.0/r-4.2.2-lv7yirk/rlib/R/library"))
    ```

1. If the prior steps fail, email research-computing@uiowa.edu with the error, asking for help to figure out how to install the packages.


## Error When Installing Packages from Source

You may have to set environment variables in each module file to help the compiler find headers and libraries.
Note that if you run into any C++ code, you will need to set the `CPLUS_INCLUDE_PATH` variable.

### `nloptr`

1. Run the following commands

```bash
module load stack/2022.1
module load r/4.1.3_gcc-9.4.0
module load nlopt
```

2. Set `LIBRARY_PATH` so linker can find library while launching `R` session (single line below):

```
LIBRARY_PATH=$ROOT_NLOPT/lib64:$LIBRARY_PATH R
```

3. In `R` console, install `nloptr` (two lines below)

```{r}
#| eval: false

install.packages(verbose = 1, "nloptr")
```

### `tkrplot`

1. Run the following commands

```bash
module load stack/2022.1
module load r/4.1.3_gcc-9.4.0
```

2. Set LIBRARY_PATH so linker can find library while launching `R` session (single line below):

```
C_INCLUDE_PATH=$ROOT_XPROTO/include LIBRARY_PATH=$ROOT_LIBXEXT/lib:$ROOT_LIBXSCRNSAVER/lib R
```

3. In `R` console, install `tkrplot` (two lines below)

```{r}
#| eval: false

install.packages(verbose = 1, "tkrplot")
```

## Error Reading from Connection

```r
Error in unserialize(node$con) : error reading from connection
Calls: parlmice ... FUN -> recvData -> recvData.SOCK0node -> unserialize
```

There likely wasn't sufficient memory for a given core.
Try increasing the max memory available and decreasing the number of cores and/or slots, so there is more memory available per core:

<https://stackoverflow.com/questions/46186375/r-parallel-error-in-unserializenodecon-in-hpc> (archived at <https://perma.cc/MF6V-NAVS>)

<https://stackoverflow.com/questions/17015598/error-calling-serialize-r-function> (archived at <https://perma.cc/3Q75-DA2D>)

<https://gforge.se/2015/02/how-to-go-parallel-in-r-basics-tips/#Memory_load> (archived at <https://perma.cc/2JRF-8Y5F>)

## Error: `$'\r': command not found`

```bash
/opt/sge/default/spool/argon-itf-bx47-36/job_scripts/5685660: line 2: $'\r': command not found
```

That often means your job script has Windows (CRLF) line endings instead of Unix (LF).
You can fix the line endings with the following command:

```bash
dos2unix your_job_file.job
```

# Appendix {#sec-appendix}

## Apptainer Definition Files {#sec-apptainerdefs}

### MPlus + R + brms + CmdStan

```bash
Bootstrap: docker
From: rocker/r-ver:latest

%files
    MplusLinux64.bin /opt/MplusLinux64.bin
    mplus_install.properties /opt/mplus_install.properties

%post
    # Install system dependencies
    apt-get update && apt-get install -y \
        libcurl4-openssl-dev \
        libssl-dev \
        libxml2-dev \
        libgit2-dev \
        libfontconfig1-dev \
        libfreetype6-dev \
        libharfbuzz-dev \
        libfribidi-dev \
        libpng-dev \
        libtiff5-dev \
        libjpeg-dev \
        libglpk-dev \
        libgmp3-dev \
        libmpfr-dev \
        make \
        g++ \
        cmake \
        git \
        sudo \
        && rm -rf /var/lib/apt/lists/*

    # Install Mplus
    chmod +x /opt/MplusLinux64.bin
    /opt/MplusLinux64.bin -i silent -f /opt/mplus_install.properties
    chmod -R 755 /opt/mplus

    # Install core CRAN packages
    R -e "install.packages(c(
        'tidyverse',
        'data.table',
        'rstan',
        'brms',
        'bayesplot',
        'mgcv',
        'lme4',
        'mirt',
        'lavaan',
        'semTools',
        'MplusAutomation',
        'SimDesign',
        'fitdistrplus',
        'sn',
        'parallelly',
        'renv',
        'remotes'
    ), repos='https://cloud.r-project.org')"

    # Install cmdstanr
    R -e "install.packages('cmdstanr', repos = c('https://stan-dev.r-universe.dev','https://cloud.r-project.org'))"

    # Install latest CmdStan into /opt/cmdstan
    mkdir -p /opt/cmdstan
    R -e "cmdstanr::install_cmdstan(cores = 2, dir='/opt/cmdstan')"

    # Enable threading for within-chain parallelization
    mkdir -p /root/.R
    echo 'STAN_THREADS=true' >> /root/.Renviron
    echo 'STAN_THREADS=true' >> /root/.R/Makevars

    # Install latest version of petersenlab package from GitHub
    R -e "remotes::install_github('DevPsyLab/petersenlab')"

%environment
    export R_LIBS_USER=/usr/local/lib/R/site-library
    export CMDSTAN=/opt/cmdstan
    export STAN_THREADS=true
    export PATH=/opt/mplus:$PATH

%runscript
    exec R "$@"
```

### R + brms + CmdStan

```bash
Bootstrap: docker
From: rocker/r-ver:latest

%post
    # Install system dependencies
    apt-get update && apt-get install -y \
        libcurl4-openssl-dev \
        libssl-dev \
        libxml2-dev \
        libgit2-dev \
        libfontconfig1-dev \
        libfreetype6-dev \
        libharfbuzz-dev \
        libfribidi-dev \
        libpng-dev \
        libtiff5-dev \
        libjpeg-dev \
        libglpk-dev \
        libgmp3-dev \
        libmpfr-dev \
        make \
        g++ \
        cmake \
        git \
        sudo \
        && rm -rf /var/lib/apt/lists/*

    # Install core CRAN packages (specify the packages to install here)
    R -e "install.packages(c(
        'tidyverse',
        'data.table',
        'rstan',
        'brms',
        'bayesplot',
        'mgcv',
        'lme4',
        'mirt',
        'lavaan',
        'semTools',
        'MplusAutomation',
        'SimDesign',
        'fitdistrplus',
        'sn',
        'parallelly',
        'renv',
        'remotes'
    ), repos='https://cloud.r-project.org')"

    # Install cmdstanr
    R -e "install.packages('cmdstanr', repos = c('https://stan-dev.r-universe.dev','https://cloud.r-project.org'))"

    # Install latest CmdStan into /opt/cmdstan
    mkdir -p /opt/cmdstan
    R -e "cmdstanr::install_cmdstan(cores = 2, dir='/opt/cmdstan')"

    # Enable threading for within-chain parallelization
    mkdir -p /root/.R
    echo 'STAN_THREADS=true' >> /root/.Renviron
    echo 'STAN_THREADS=true' >> /root/.R/Makevars

    # Install latest version of petersenlab package from GitHub
    R -e "remotes::install_github('DevPsyLab/petersenlab')"

%environment
    export R_LIBS_USER=/usr/local/lib/R/site-library
    export CMDSTAN=/opt/cmdstan
    export STAN_THREADS=true

%runscript
    exec R "$@"
```
    
## Old Stacks

### Stack 2022.1

```bash
module load stack/2022.1
module load r/4.1.3_gcc-9.4.0
cd path
```

### Stack 2021.1

```bash
module load stack/2021.1
module load r/4.0.5_gcc-9.3.0
cd path
```

### Stack 2020.2

```bash
module load stack/2020.2
module load r/4.0.2_gcc-8.4.0
cd path
```

### Stack 2020.1

```bash
module load stack/2020.1
module load r/3.6.2_gcc-9.2.0
cd path
```

## Managing packages for a project using `packrat`

Please note: we now use [`renv`](#sec-renv) rather than `packrat` for package management

<https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76514803/Adding+R+programs+to+a+personal+library> (archived at <https://perma.cc/3SRR-2JE7>)

1. Create a directory where you want you want to store the project.
1. `cd` to the above directory
1. Start R
1. Load the `packrat` package: `library("packrat")`
1. At the `R` prompt, initialize the `packrat` project on the local repository of `R` packages with: `packrat::init(project = "/Users/itpetersen/Documents/Projects/INSERT_PROJECT_NAME/", options = list(local.repos = "/Users/itpetersen/R/x86_64-pc-linux-gnu-library/4.0"))`
1. You must restart your `R` session in the given project directory after running init in order for the changes to take effect!
1. From this point on you are working in a `packrat` project.
Installed packages will go into a library within this project.
After initializing the `packrat` project on the local repository of `R` packages, packages from the local repository can be installed with `packrat::install_local()`:
    - `packrat::install_local("package_name")`
1. To restart a packrat project simply start `R` from the directory created in step (1). The project will initialize automatically.

To save the current state of your library: `packrat::snapshot()`; if that command fails due to an error when fetching sources, try `packrat::snapshot(snapshot.sources = FALSE)` 

To disable `packrat` on a project: `disable(restart = FALSE)`

### To install a new package for a project after `packrat` has been initialized

1. First, load the `R` environment module (see `Install R packages` section above)
1. cd to the directory of the `R` project
1. launch R
    - `R`
1. then install the package, (the code below uses a generic `package_name`):
    - `install.packages("package_name", repos = "http://cran.r-project.org", type = "source", dependencies = TRUE, Ncpus = 40)`
    - `install.packages(c("packrat","psych","tidyverse","data.table","nlme","lme4","mirt","TeachingDemos","Amelia","mice","miceadds","abind","future","lavaan","blavaan","Rcpp","igraph","shinystan","StanHeaders","brms","rstan","rjags","renv"), repos = "http://cran.r-project.org", type = "source", dependencies = TRUE, Ncpus = 40)`
1. `packrat::snapshot()`; if that command fails due to an error when fetching sources, try `packrat::snapshot(snapshot.sources = FALSE)`

### Build `packrat` environment on a compute node (only if necessary if package load fails due to issues building packages)

1. `qlogin` -- for more info, see here: <https://uiowa.atlassian.net/wiki/spaces/hpcdocs/pages/76513454/Qlogin+for+Interactive+Sessions> (archived at <https://perma.cc/Y3A9-WQ3W>)
1. `module load stack/2020.2-base_arch` -- this will ensure that the modules point to the lowest common multiarchitecture on Argon and will run on all nodes
1. `module load r`
1. cd to the directory of the `R` project
1. `R`
1. Create `packrat` project (see above)
1. Install packages
1. Take a package snapshot using `packrat`
1. `exit` -- to exit the qlogin session

Note: Your `packrat` environment will then be linked to the proper `glpk` library and will run on any Argon node.
You do not need to use the `2020.2-base_arch` module at run time, only build time.
