---
title: "Mplus"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  comment = "",
  class.source = "fold-show")
```

# Overview of `Mplus`

[`Mplus`](https://www.statmodel.com) is software for [structural equation modeling](sem.html).

# Model Estimators {#estimators}

- `MLR`: for likert/continuous data
- `WLSMV`: for ordinal/categorical data
- `bayes`

# Parameter Label {#parameterLabel}

To specify a parameter label, provide the label in parenthese after the parameter:

```
MODEL:
    y1 BY x1* x2 x3 (load1-3);
    y2 BY x4* x5 (load5) x6;
```

# Cluster Variable {#clusterVariable}

There are multiple ways of accounting for nested data in [structural equation modeling](sem.html).
One way to account for nested data is to use multilevel structural equation modeling.
Another approach is to use a cluster variable to generate cluster-robust standard errors of parameters.
To use a cluster variable, specify `CLUSTER` under the `VARIABLE` section, and specify `TYPE = COMPLEX` under the `ANALYSIS` section:

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CLUSTER = ID;

ANALYSIS:
    TYPE = COMPLEX;
```

# Low Covariance Coverage {#lowCovarianceCoverage}

To estimate a model with low covariance coverage, lower the `COVERAGE` value under the `ANALYSIS` section:

```
ANALYSIS:
    COVERAGE = 0;
```

# Setting Parameter Constraints {#parameterConstraints}

## Setting Two Parameters to be Equal {#equalParameters}

To set two parameters to be equal, provide the same [parameter label](#parameterLabel) for each parameter.

## Setting Lower and Upper Bounds on a Parameter {#parameterBounds}

To set lower and upper bounds on a parameter, you can assign the parameter a [parameter label](#parameterLabel).
Then, you can assign the constraint to the parameter (via the label) under the `MODEL CONSTRAINT` section.
For example, to constrain a parameter between 0–1, 

```
MODEL:
    y1 BY x1* x2 x3 (load3);

MODEL CONSTRAINT:
    load3 > 0; load3 < 1;
```

# Session Info

```{r, class.source = "fold-hide"}
sessionInfo()
```
