---
title: "Mplus"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  comment = "",
  class.source = "fold-show")
```

# Overview of `Mplus`

[`Mplus`](https://www.statmodel.com) is software for [structural equation modeling](sem.html).
A summary of the [`Mplus`](https://www.statmodel.com) language syntax is [here](https://www.statmodel.com/language.html) (archived at: https://perma.cc/962G-QUEG).
The [Mplus User's Guide](https://www.statmodel.com/html_ug.shtml) is located [here](https://www.statmodel.com/download/usersguide/MplusUserGuideVer_8.pdf) (archived at: https://perma.cc/W39W-NRGH).

# Prepare Data {#prepareData}

To prepare the data in `R` for use in `Mplus`:

```{r, eval = FALSE}
library("MplusAutomation")

prepareMplusData(
  mydata,
  file.path(path, "/GitHub/Project_Name/Data/mplusdata.dat"))
```

# Model Example {#example}

```
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
!!!!!  MPLUS SYNTAX LINES CANNOT EXCEED 90 CHARACTERS;
!!!!!  VARIABLE NAMES AND PARAMETER LABELS CANNOT EXCEED 8 CHARACTERS EACH;
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

TITLE: Example Mplus Model

DATA: 
    FILE = "C:/[insert_filepath]/filename.dat";

VARIABLE:
    NAMES = ID age x1 x2 x3 x4 x5 x6 y1 y2;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 x3 x4 x5 x6 y1 y2;

ANALYSIS:
    TYPE = COMPLEX;
    ESTIMATOR = MLR;

MODEL:
    ! Factor loadings
    latent1 BY x1* x2 x3;
    latent2 BY x4* x5 x6;

    ! Covariances between latent factors
    latent1 WITH latent2

    ! Regression paths
    y1 ON latent1 + latent2
    y2 ON latent1 + latent 2

    ! Standardize latent factors: fix means to zero
    [latent1@0];
    [latent2@0];

    ! Standardize latent factors: fix variances to one
    latent1@1;
    latent2@1;

OUTPUT:  
    STDYX; 
    TECH1;
    TECH4;
    SAMPSTAT;
    MODINDICES (3);
    CINTERVAL;

SAVEDATA:
    FILE = "C:/[insert_filepath]/filename.dat";
    SAVE = FSCORES;
```

# Model Title {#modelTitle}

```
TITLE: INSERT TITLE HERE
```

# Read Data {#readData}

```
DATA: 
    FILE = "C:/[insert_filepath]/filename.dat";
```

# Variables {#variables}

## Specify Variables {#specifyVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CLUSTER = ID;
```

## Categorical Variables {#categoricalVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CATEGORICAL = x1 x2;
```

## Count Variables {#countVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    COUNT = x1 x2;
```

## Cluster Variable {#clusterVariable}

There are multiple ways of accounting for nested data in [structural equation modeling](sem.html).
One way to account for nested data is to use multilevel structural equation modeling.
Another approach is to use a cluster variable to generate cluster-robust standard errors of parameters.
To use a cluster variable, specify `CLUSTER` under the `VARIABLE` section, and specify `TYPE = COMPLEX` under the `ANALYSIS` section:

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CLUSTER = ID;

ANALYSIS:
    TYPE = COMPLEX;
```

## Individually Varying Times of Observation {#tScores}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    TSCORES = age;
```

## Auxiliary Variables {#auxiliaryVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID x1 x2 y1;
    AUXILIARY = age;
```

## Sampling Weight Variable {#samplingWeight}

```
VARIABLE:
    NAMES = ID wt age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    WEIGHT = wt;
```

## Multilevel Variables {#multilevelVariables}

Between- and within-cluster variables:

```
VARIABLE:
    NAMES = ID wt age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    WITHIN = x1;
    BETWEEN = x2;
```

# Analysis {#analysis}

## Analysis Types {#analysisTypes}

- `TYPE = COMPLEX`
- `TYPE = TWOLEVEL`
- `TYPE = EFA`

## Model Estimators {#estimators}

```
ANALYSIS:
    ESTIMATOR = MLR;
```

- `MLR`: for likert/continuous data
- `WLSMV`: for ordinal/categorical data
- `BAYES`

## Bootstrap Draws {#bootstrap}

```
BOOTSTRAP = 2000; ! insert number of bootstrap draws
```

## Starts {#starts}

```
STARTS = 20; ! insert number of initial stage starts and number of final stage optimizations
```

## Low Covariance Coverage {#lowCovarianceCoverage}

To estimate a model with low covariance coverage, lower the `COVERAGE` value under the `ANALYSIS` section:

```
ANALYSIS:
    COVERAGE = 0;
```

# Exploratory Factor Analysis {#efa}

## Oblique Rotation {#oblique}

```
ANALYSIS:
    TYPE = EFA 1 5; ! extract 1-5 factors
    ROTATION = GEOMIN;
```

## Orthogonal Rotation {#orthogonal}

```
ANALYSIS:
    TYPE = EFA 1 5; ! extract 1-5 factors
    ROTATION = VARIMAX;
```

# Bayesian SEM {#bayesian}

Other settings you can specify under the "ANALYSIS" section include:

```
ANALYSIS:
    ESTIMATOR = BAYES;
    BCONVERGENCE = .05; ! value of the Gelman-Rubin convergence criterion; ! default is .05; van de Schoot et al. (2014) recommend .01
    BITERATIONS = a (b); ! a = maximum, b = minumum number of iterations for each MCMC chain
    CHAIN = 4; ! number of chains
    PROCESSORS = 4; ! number of computer processors to use
    BSEED = 52242; ! set seed for replicability
    STVALUES = ml; ! set starting values based on ML estimation
```

You can specify model priors under the "MODEL PRIORS" section.

Other settings you can specify under the "OUTPUT" section include:

```
OUTPUT:
    STAND; ! standardized estimates
    CINTERCAL; ! posterior predictive intervals (credible intervals)
```

Other settings you can specify under the "PLOT" section include:

```
PLOT:
    TYPE = PLOT3; ! trace plots, histogram, and kernel density
```

# Model {#model}

## Define Latent Variables {#latentVariables}

```
MODEL:
    latent1 BY x1 x2 x3;
```

## Regression Paths {#regressionPaths}

Regress outcome variable on predictor variable(s):

```
MODEL:
    y1 ON x1 x2;
```

## Covariance Paths {#covariancePaths}

```
MODEL:
    x1 WITH x2;
```

## Indirect Effects {#indirectEffects}

```
MODEL INDIRECT:
    y IND x;
```

## Means/Intercepts {#means}

Freely estimate:

```
MODEL:
    [x1];
```

Fix to zero:

```
MODEL:
    [x1@0];
```

## Variances {#variances}

Freely estimate:

```
MODEL:
    x1;
```

Fix to one:

```
MODEL:
    x1@1;
```

## Parameter Label {#parameterLabel}

To specify a parameter label, provide the label in parentheses after the parameter:

```
MODEL:
    latent1 BY x1* x2 x3 (load1-3);
    latent2 BY x4* x5 (load5) x6;
```

# Comments {#comments}

```
!This is a comment in Mplus
```

# Setting Parameter Constraints {#parameterConstraints}

## Freeing a Parameter {#freeParameter}

By default, the first loading on a factor is fixed to zero.
You can freely estimate the parameter by adding an asterisk:

```
MODEL:
    latent1 BY x1* x2 x3;
```

## Constraing a Parameter {#constrainParameter}

```
MODEL:
    latent1 BY x1@1 x2 x3;
    [latent1@0];
    latent1@1;
```

## Setting Two Parameters to be Equal {#equalParameters}

To set two parameters to be equal, provide the same [parameter label](#parameterLabel) for each parameter.

## Setting Lower and Upper Bounds on a Parameter {#parameterBounds}

To set lower and upper bounds on a parameter, you can assign the parameter a [parameter label](#parameterLabel).
Then, you can assign the constraint to the parameter (via the label) under the `MODEL CONSTRAINT` section.
For example, to constrain a parameter between 0–1, 

```
MODEL:
    latent1 BY x1* x2 x3 (load3);

MODEL CONSTRAINT:
    load3 > 0; load3 < 1;
```

# Output {#output}

```
OUTPUT:  
    STDYX; 
    TECH1;
    TECH4;
    SAMPSTAT;
    MODINDICES (ALL); ! specify number in parentheses to print only those mod indices that are above a particular chi-square value
    CINTERVAL;
```

# Save Factor Scores {#factorScores}

```
SAVEDATA:
    FILE = "C:/[insert_filepath]/filename.dat";
    SAVE = FSCORES;
```

# Multiple Imputation {#multipleImputation}

For examples of how to conduct [multiple imputation](multipleImputation.html) in [`Mplus`](https://www.statmodel.com), see [here](multipleImputation.html#mplus).

# Run Models in Batch {#runModels}

Run models in batch via `R`:

```{r, eval = FALSE}
library("MplusAutomation")

runModels(
  file.path(path, "/GitHub/Project_Name/Analyses/Mplus/"),
  recursive = TRUE,
  replaceOutfile = "always",
  showOutput = TRUE)
```

# Session Info

```{r, class.source = "fold-hide"}
sessionInfo()
```
