---
title: "Mplus"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  comment = "",
  class.source = "fold-show")
```

# Overview of `Mplus`

[`Mplus`](https://www.statmodel.com) is software for [structural equation modeling](sem.html).
A summary of the [`Mplus`](https://www.statmodel.com) language syntax is [here](https://www.statmodel.com/language.html) (archived at: https://perma.cc/962G-QUEG).
The [Mplus User's Guide](https://www.statmodel.com/html_ug.shtml) is located [here](https://www.statmodel.com/download/usersguide/MplusUserGuideVer_8.pdf) (archived at: https://perma.cc/W39W-NRGH).

# Prepare Data {#prepareData}

To prepare the data in `R` for use in `Mplus`:

```{r, eval = FALSE}
library("MplusAutomation")

prepareMplusData(
  mydata,
  file.path(path, "/GitHub/Project_Name/Data/mplusdata.dat"))
```

# Model Example {#example}

```
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
!!!!!  MPLUS SYNTAX LINES CANNOT EXCEED 90 CHARACTERS;
!!!!!  VARIABLE NAMES AND PARAMETER LABELS CANNOT EXCEED 8 CHARACTERS EACH;
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

TITLE: Example Mplus Model

DATA: 
    FILE = "C:/[insert_filepath]/filename.dat";

VARIABLE:
    NAMES = ID age x1 x2 x3 x4 x5 x6 y1 y2;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 x3 x4 x5 x6 y1 y2;

ANALYSIS:
    TYPE = COMPLEX;
    ESTIMATOR = MLR;

MODEL:
    ! Factor loadings
    latent1 BY x1* x2 x3;
    latent2 BY x4* x5 x6;

    ! Covariances between latent factors
    latent1 WITH latent2

    ! Regression paths
    y1 ON latent1 + latent2
    y2 ON latent1 + latent 2

    ! Standardize latent factors: fix means to zero
    [latent1@0];
    [latent2@0];

    ! Standardize latent factors: fix variances to one
    latent1@1;
    latent2@1;

OUTPUT:  
    STDYX; 
    TECH1;
    TECH4;
    SAMPSTAT;
    MODINDICES (3);
    CINTERVAL;
    RESIDUAL;

SAVEDATA:
    FILE = "C:/[insert_filepath]/filename.dat";
    SAVE = FSCORES;
```

# Model Title {#modelTitle}

```
TITLE: INSERT TITLE HERE
```

# Read Data {#readData}

```
DATA: 
    FILE = "C:/[insert_filepath]/filename.dat";
```

# Variables {#variables}

## Specify Variables {#specifyVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CLUSTER = ID;
```

## Categorical Variables {#categoricalVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CATEGORICAL = x1 x2;
```

## Count Variables {#countVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    COUNT = x1 x2;
```

## Cluster Variable {#clusterVariable}

There are multiple ways of accounting for nested data in [structural equation modeling](sem.html).
One way to account for nested data is to use multilevel structural equation modeling.
Another approach is to use a cluster variable to generate cluster-robust standard errors of parameters.
To use a cluster variable, specify `CLUSTER` under the `VARIABLE` section, and specify `TYPE = COMPLEX` under the `ANALYSIS` section:

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    CLUSTER = ID;

ANALYSIS:
    TYPE = COMPLEX;
```

## Individually Varying Times of Observation {#tScores}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    TSCORES = age;
```

## Auxiliary Variables {#auxiliaryVariables}

```
VARIABLE:
    NAMES = ID age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID x1 x2 y1;
    AUXILIARY = age;
```

## Sampling Weight Variable {#samplingWeight}

```
VARIABLE:
    NAMES = ID wt age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    WEIGHT = wt;
```

## Multilevel Variables {#multilevelVariables}

Between- and within-cluster variables:

```
VARIABLE:
    NAMES = ID wt age x1 x2 x3 y1;
    MISSING = .;
    USEVARIABLES = ID age x1 x2 y1;
    WITHIN = x1;
    BETWEEN = x2;
```

# Analysis {#analysis}

## Analysis Types {#analysisTypes}

- `TYPE = COMPLEX`
- `TYPE = TWOLEVEL`
- `TYPE = EFA`

## Model Estimators {#estimators}

```
ANALYSIS:
    ESTIMATOR = MLR;
```

- `MLR`: for likert/continuous data
- `WLSMV`: for ordinal/categorical data
- `BAYES`

## Bootstrap Draws {#bootstrap}

```
BOOTSTRAP = 2000; ! insert number of bootstrap draws
```

## Starts {#starts}

```
STARTS = 20; ! insert number of initial stage starts and number of final stage optimizations
```

## Low Covariance Coverage {#lowCovarianceCoverage}

To estimate a model with low covariance coverage, lower the `COVERAGE` value under the `ANALYSIS` section:

```
ANALYSIS:
    COVERAGE = 0;
```

# Exploratory Factor Analysis {#efa}

## Oblique Rotation {#oblique}

```
ANALYSIS:
    TYPE = EFA 1 5; ! extract 1-5 factors
    ROTATION = GEOMIN;
```

## Orthogonal Rotation {#orthogonal}

```
ANALYSIS:
    TYPE = EFA 1 5; ! extract 1-5 factors
    ROTATION = VARIMAX;
```

# Bayesian SEM {#bayesian}

Other settings you can specify under the "ANALYSIS" section include:

```
ANALYSIS:
    ESTIMATOR = BAYES;
    BCONVERGENCE = .05; ! value of the Gelman-Rubin convergence criterion; ! default is .05; van de Schoot et al. (2014) recommend .01
    BITERATIONS = a (b); ! a = maximum, b = minumum number of iterations for each MCMC chain
    CHAIN = 4; ! number of chains
    PROCESSORS = 4; ! number of computer processors to use
    BSEED = 52242; ! set seed for replicability
    STVALUES = ml; ! set starting values based on ML estimation
```

You can specify model priors under the "MODEL PRIORS" section.

Other settings you can specify under the "OUTPUT" section include:

```
OUTPUT:
    STAND; ! standardized estimates
    TECH1; ! model priors
    TECH8; ! potential scale reduction (PSR); to evaluate convergence (should be near one)
    CINTERVAL; ! posterior predictive intervals (credible intervals)
```

Other settings you can specify under the "PLOT" section include:

```
PLOT:
    TYPE = PLOT3; ! trace plots, histogram, and kernel density
```

# Model {#model}

## Define Latent Variables {#latentVariables}

```
MODEL:
    latent1 BY x1 x2 x3;
```

## Regression Paths {#regressionPaths}

Regress outcome variable on predictor variable(s):

```
MODEL:
    y1 ON x1 x2;
```

## Covariance Paths {#covariancePaths}

```
MODEL:
    x1 WITH x2;
```

## Indirect Effects {#indirectEffects}

```
ANALYSIS:
    TYPE = GENERAL;
    ESTIMATOR = ML;
    BOOTSTRAP = 1000;

MODEL:
    MODEL INDIRECT:
        y IND x;

OUTPUT:
    STAND;
    CINTERVAL (BOOTSTRAP); !percentile boostrap CI
    CINTERVAL (BCBOOTSTRAP); !bias-corrected boostrap CI
```

## Means/Intercepts {#means}

Freely estimate:

```
MODEL:
    [x1];
```

Fix to zero:

```
MODEL:
    [x1@0];
```

## Variances {#variances}

Freely estimate:

```
MODEL:
    x1;
```

Fix to one:

```
MODEL:
    x1@1;
```

## Parameter Label {#parameterLabel}

To specify a parameter label, provide the label in parentheses after the parameter:

```
MODEL:
    latent1 BY x1* x2 x3 (load1-3);
    latent2 BY x4* x5 (load5) x6;
```

## Multigroup Model {#multigroup}

```
VARIABLE:
    NAMES = group x1 x2 x3 y1;
    GROUPING = group (0=boys, 1=girls);
    MISSING = .;
    USEVARIABLES = group x1 x2 x3 y1;

MODEL:
    Model boys:
        latent BY x1* x2 x3;
        [latent@0];
        latent@1;
        y ~ latent;

    Model girls:
        latent BY x1* x2 x3;
        [latent@0];
        latent@1;
        y ~ latent;
```

# Comments {#comments}

```
!This is a comment in Mplus
```

# Setting Parameter Constraints {#parameterConstraints}

## Freeing a Parameter {#freeParameter}

By default, the first loading on a factor is fixed to zero.
You can freely estimate the parameter by adding an asterisk:

```
MODEL:
    latent1 BY x1* x2 x3;
```

## Constraing a Parameter {#constrainParameter}

```
MODEL:
    latent1 BY x1@1 x2 x3;
    [latent1@0];
    latent1@1;
```

## Setting Two Parameters to be Equal {#equalParameters}

To set two parameters to be equal, provide the same [parameter label](#parameterLabel) for each parameter.

## Setting Lower and Upper Bounds on a Parameter {#parameterBounds}

To set lower and upper bounds on a parameter, you can assign the parameter a [parameter label](#parameterLabel).
Then, you can assign the constraint to the parameter (via the label) under the `MODEL CONSTRAINT` section.
For example, to constrain a parameter between 0–1, 

```
MODEL:
    latent1 BY x1* x2 x3 (load3);

MODEL CONSTRAINT:
    load3 > 0; load3 < 1;
```

# Output {#output}

```
OUTPUT:  
    STDYX; 
    TECH1;
    TECH4;
    SAMPSTAT;
    MODINDICES (ALL); ! specify number in parentheses to print only those mod indices that are above a particular chi-square value
    CINTERVAL;
    RESIDUAL;
```

# Save Factor Scores {#factorScores}

```
SAVEDATA:
    FILE = "C:/[insert_filepath]/filename.dat";
    SAVE = FSCORES;
```

# Multiple Imputation {#multipleImputation}

For examples of how to conduct [multiple imputation](multipleImputation.html) in [`Mplus`](https://www.statmodel.com), see [here](multipleImputation.html#mplus).
To run a model on multiply imputed data, use the following:

```
DATA: FILE = "implist.dat"; ! where implist.dat is the name of the *list.dat file saved from the imputation step
    TYPE = IMPUTATION;
```

# Run Models in Batch {#runModels}

Run models in batch via `R`:

```{r, eval = FALSE}
library("MplusAutomation")

runModels(
  file.path(path, "/GitHub/Project_Name/Analyses/Mplus/"),
  recursive = TRUE,
  replaceOutfile = "always",
  showOutput = TRUE)
```

# Monte Carlo Simulation/Power Analysis {#monteCarlo}

## Single Group Model {#monteCarloSingleGroup}

```
TITLE: Single-Group Monte Carlo Simulation with Ordinal Items and Common Factor;

MONTECARLO:
    NAMES = v1-v5; ! variable names
    NOBSERVATIONS = 500; ! number of participants in each sample
    NREPS = 100000; ! number of samples to create
    SEED = 52242; ! random seed
    GENERATE = v1-v5 (3 p); ! specify the scale of the DVs; number of thresholds; probit (p)
    CATEGORICAL = v1-v5; ! specify the variables that are (ordered) categorical

ANALYSIS:
    PROCESSORS = 4 1; ! number of processors; number of threads
    ESTIMATOR = WLSMV;
    PARAMETERIZATION = THETA;

MODEL POPULATION: ! tell Mplus how to generate the population data; can use asterisks (*) and at symbols (@) interchangeably here, but they differ in the MODEL command (see below); I use the same symbol as in the MODEL command
    dep BY v1-v5*.7; ! factor loadings
    [dep@0]; ! set factor mean to 0
    dep@1; ! set factor variance to 1 (standardize)
    [v1$1*0.5] (v1t1); [v1$2*1.0] (v1t2); [v1$3*1.5] (v1t3); ! item thresholds for v1
    [v2$1*0.5] (v2t1); [v2$2*1.0] (v2t2); [v2$3*1.5] (v2t3); ! item thresholds for v2
    [v3$1*0.5] (v3t1); [v3$2*1.0] (v3t2); [v3$3*1.5] (v3t3); ! item thresholds for v3
    [v4$1*0.0] (v4t1); [v4$2*0.5] (v4t2); [v4$3*1.0] (v4t3); ! item thresholds for v4
    [v5$1*0.0] (v5t1); [v5$2*0.5] (v5t2); [v5$3*1.0] (v5t3); ! item thresholds for v5
    v1-v5@1; ! item residual variances

MODEL: ! tell Mplus to estimate our model; asterisks (*) are free estimates with a starting value; at symbols (@) are fixed estimates
    dep BY v1-v5*.7; ! factor loadings
    [dep@0]; ! set factor mean to 0
    dep@1; ! set factor variance to 1 (standardize)
    [v1$1*0.5] (v1t1); [v1$2*1.0] (v1t2); [v1$3*1.5] (v1t3); ! item thresholds for v1
    [v2$1*0.5] (v2t1); [v2$2*1.0] (v2t2); [v2$3*1.5] (v2t3); ! item thresholds for v2
    [v3$1*0.5] (v3t1); [v3$2*1.0] (v3t2); [v3$3*1.5] (v3t3); ! item thresholds for v3
    [v4$1*0.0] (v4t1); [v4$2*0.5] (v4t2); [v4$3*1.0] (v4t3); ! item thresholds for v4
    [v5$1*0.0] (v5t1); [v5$2*0.5] (v5t2); [v5$3*1.0] (v5t3); ! item thresholds for v5
    v1-v5@1; ! item residual variances

MODEL CONSTRAINT:
    NEW (stdt stdt1 stdt2 stdt3 noninvt noninvt1 noninvt2 noninvt3 diff);
    
    stdt1 = (v1t1 + v2t1 + v3t1) / 3;
    stdt2 = (v1t2 + v2t2 + v3t2) / 3;
    stdt3 = (v1t3 + v2t3 + v3t3) / 3;
    
    noninvt1 = (v4t1 + v5t1) / 2;
    noninvt2 = (v4t2 + v5t2) / 2;
    noninvt3 = (v4t3 + v5t3) / 2;
    
    stdt = (stdt1 + stdt2 + stdt3) / 3;
    noninvt = (noninvt1 + noninvt2 + noninvt3) / 3;
    
    diff = noninvt - stdt;

OUTPUT:
    TECH9;
```

## Multi-Group Model {#monteCarloMultigroup}

```
TITLE: Multi-Group Monte Carlo Simulation with Ordinal Items and Common Factor;

MONTECARLO:
    NAMES = v1-v5; ! variable names
    NGROUPS = 2; ! number of groups
    NOBSERVATIONS = 500 300; ! number of participants in each sample
    NREPS = 100000; ! number of samples to create
    SEED = 52242; ! random seed
    GENERATE = v1-v5 (3 p); ! specify the scale of the DVs; number of thresholds; probit (p)
    CATEGORICAL = v1-v5; ! specify the variables that are (ordered) categorical

ANALYSIS:
    PROCESSORS = 4 1; ! number of processors; number of threads
    ESTIMATOR = WLSMV;
    PARAMETERIZATION = THETA;

MODEL POPULATION: ! tell Mplus how to generate the population data; can use asterisks (*) and at symbols (@) interchangeably here, but they differ in the MODEL command (see below); I use the same symbol as in the MODEL command
    dep BY v1-v5*.7; ! factor loadings
    [dep@0]; ! set factor mean to 0
    dep@1; ! set factor variance to 1 (standardize)
    [v1$1*0.5] (v1t1g1); [v1$2*1.0] (v1t2g1); [v1$3*1.5] (v1t3g1); ! item thresholds for v1
    [v2$1*0.5] (v2t1g1); [v2$2*1.0] (v2t2g1); [v2$3*1.5] (v2t3g1); ! item thresholds for v2
    [v3$1*0.5] (v3t1g1); [v3$2*1.0] (v3t2g1); [v3$3*1.5] (v3t3g1); ! item thresholds for v3
    [v4$1*0.5] (v4t1g1); [v4$2*1.0] (v4t2g1); [v4$3*1.5] (v4t3g1); ! item thresholds for v4
    [v5$1*0.5] (v5t1g1); [v5$2*1.0] (v5t2g1); [v5$3*1.5] (v5t3g1); ! item thresholds for v5
    v1-v5@1; ! item residual variances
    
MODEL POPULATION-g2: ! tell Mplus how to generate the population data for group 2; can use asterisks (*) and at symbols (@) interchangeably here, but they differ in the MODEL command (see below); I use the same symbol as in the MODEL command
    !dep BY v1-v5*.7; ! factor loadings
    ![dep@0]; ! set factor mean to 0
    !dep@1; ! set factor variance to 1 (standardize)
    [v1$1*0.5] (v1t1g2); [v1$2*1.0] (v1t2g2); [v1$3*1.5] (v1t3g2); ! item thresholds for v1
    [v2$1*0.5] (v2t1g2); [v2$2*1.0] (v2t2g2); [v2$3*1.5] (v2t3g2); ! item thresholds for v2
    [v3$1*0.5] (v3t1g2); [v3$2*1.0] (v3t2g2); [v3$3*1.5] (v3t3g2); ! item thresholds for v3
    [v4$1*0.0] (v4t1g2); [v4$2*0.5] (v4t2g2); [v4$3*1.0] (v4t3g2); ! item thresholds for v4
    [v5$1*0.0] (v5t1g2); [v5$2*0.5] (v5t2g2); [v5$3*1.0] (v5t3g2); ! item thresholds for v5
    !v1-v5@1; ! item residual variances

MODEL: ! tell Mplus to estimate our model; asterisks (*) are free estimates with a starting value; at symbols (@) are fixed estimates
    dep BY v1-v5*.7; ! factor loadings
    [dep@0]; ! set factor mean to 0
    dep@1; ! set factor variance to 1 (standardize)
    [v1$1*0.5] (v1t1g1); [v1$2*1.0] (v1t2g1); [v1$3*1.5] (v1t3g1); ! item thresholds for v1
    [v2$1*0.5] (v2t1g1); [v2$2*1.0] (v2t2g1); [v2$3*1.5] (v2t3g1); ! item thresholds for v2
    [v3$1*0.5] (v3t1g1); [v3$2*1.0] (v3t2g1); [v3$3*1.5] (v3t3g1); ! item thresholds for v3
    [v4$1*0.5] (v4t1g1); [v4$2*1.0] (v4t2g1); [v4$3*1.5] (v4t3g1); ! item thresholds for v4
    [v5$1*0.5] (v5t1g1); [v5$2*1.0] (v5t2g1); [v5$3*1.5] (v5t3g1); ! item thresholds for v5
    v1-v5@1; ! item residual variances
   
MODEL g2: ! tell Mplus to estimate our model in group 2; asterisks (*) are free estimates with a starting value; at symbols (@) are fixed estimates
    !dep BY v1-v5*.7; ! factor loadings
    ![dep@0]; ! set factor mean to 0
    !dep@1; ! set factor variance to 1 (standardize)
    [v1$1*0.5] (v1t1g2); [v1$2*1.0] (v1t2g2); [v1$3*1.5] (v1t3g2); ! item thresholds for v1
    [v2$1*0.5] (v2t1g2); [v2$2*1.0] (v2t2g2); [v2$3*1.5] (v2t3g2); ! item thresholds for v2
    [v3$1*0.5] (v3t1g2); [v3$2*1.0] (v3t2g2); [v3$3*1.5] (v3t3g2); ! item thresholds for v3
    [v4$1*0.0] (v4t1g2); [v4$2*0.5] (v4t2g2); [v4$3*1.0] (v4t3g2); ! item thresholds for v4
    [v5$1*0.0] (v5t1g2); [v5$2*0.5] (v5t2g2); [v5$3*1.0] (v5t3g2); ! item thresholds for v5
    !v1-v5@1; ! item residual variances

MODEL CONSTRAINT:
    NEW (stdt1g1 stdt2g1 stdt3g1 stdt1g2 stdt2g2 stdt3g2
    nonit1g1 nonit2g1 nonit3g1 nonit1g2 nonit2g2 nonit3g2
    stdtg1 stdtg2 nonitg1 nonitg2 diffwg diffbg);
    
    stdt1g1 = (v1t1g1 + v2t1g1 + v3t1g1) / 3;
    stdt2g1 = (v1t2g1 + v2t2g1 + v3t2g1) / 3;
    stdt3g1 = (v1t3g1 + v2t3g1 + v3t3g1) / 3;
    
    stdt1g2 = (v1t1g2 + v2t1g2 + v3t1g2) / 3;
    stdt2g2 = (v1t2g2 + v2t2g2 + v3t2g2) / 3;
    stdt3g2 = (v1t3g2 + v2t3g2 + v3t3g2) / 3;
    
    nonit1g1 = (v4t1g1 + v5t1g1) / 2;
    nonit2g1 = (v4t2g1 + v5t2g1) / 2;
    nonit3g1 = (v4t3g1 + v5t3g1) / 2;
    
    nonit1g2 = (v4t1g2 + v5t1g2) / 2;
    nonit2g2 = (v4t2g2 + v5t2g2) / 2;
    nonit3g2 = (v4t3g2 + v5t3g2) / 2;
    
    stdtg1 = (stdt1g1 + stdt2g1 + stdt3g1) / 3;
    stdtg2 = (stdt1g2 + stdt2g2 + stdt3g2) / 3;
    
    nonitg1 = (nonit1g1 + nonit2g1 + nonit3g1) / 3;
    nonitg2 = (nonit1g2 + nonit2g2 + nonit3g2) / 3;
    
    diffwg = nonitg2 - stdtg2; ! difference within group
    
    diffbg = nonitg2 - nonitg1; ! difference between groups
    
OUTPUT:
    TECH9;
```

# Session Info

```{r, class.source = "fold-hide"}
sessionInfo()
```
