---
title: "Multiple Imputation"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      comment = "",
                      class.source = "fold-show")
```

# Multilevel Multiple Imputation

## Three-Level and Cross-Classified Data

https://simongrund1.github.io/posts/multiple-imputation-for-three-level-and-cross-classified-data/

## Methods

### Continuous Outcomes

https://stefvanbuuren.name/fimd/sec-multioutcome.html#methods

```{r, eval = FALSE}
?mice::mice.impute.2l.norm
?mice::mice.impute.2l.pan
?mice::mice.impute.2l.lmer
?miceadds::mice.impute.2l.pmm
?miceadds::mice.impute.2l.contextual.pmm
?miceadds::mice.impute.2l.continuous
?micemd::mice.impute.2l.2stage.norm
?micemd::mice.impute.2l.2stage.pmm
?micemd::mice.impute.2l.glm.norm
?micemd::mice.impute.2l.jomo
```

### Discrete Outcomes

https://stefvanbuuren.name/fimd/sec-catoutcome.html#methods-1

```{r, eval = FALSE}
?mice::mice.impute.2l.bin
?miceadds::mice.impute.2l.binary
?micemd::mice.impute.2l.2stage.bin
?micemd::mice.impute.2l.glm.bin
```

## Specifying the Predictor Matrix

- cluster variable: `-2`
- fixed effect of predictor: `1`
- fixed effect and random effect of predictor: `2`
- include cluster mean of predictor in addition to fixed effect of predictor: `3`
- include cluster mean of predictor in addition to fixed effect and random effect of predictor: `4`

# Parellel Processing

https://www.gerkovink.com/miceVignettes/futuremice/Vignette_futuremice.html

# Imputation Diagnostics

## Trace Plots

On convergence, the streams of the trace plots should intermingle and be free of any trend (at the later iterations).
Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence.

- https://stefvanbuuren.name/fimd/sec-algoptions.html

```{r, eval = FALSE}
plot(mice_object, c("variable"))
```

## Density Plots

```{r, eval = FALSE}
densityplot(mice_object)
densityplot(mice_object, ~ variable)
```

## Strip Plots

```{r, eval = FALSE}
stripplot(mice_object)
stripplot(mice_object, variable ~ .imp)
```

# Resources

https://stefvanbuuren.name/fimd

https://www.gerkovink.com/miceVignettes/Multi_level/Multi_level_data.html
