---
title: "Multiple Imputation"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      comment = "",
                      class.source = "fold-show")
```

# Preamble

## Install Libraries

```{r}
#install.packages("remotes")
#remotes::install_git("https://research-git.uiowa.edu/PetersenLab/petersenlab.git")
```

## Load Libraries

```{r, message = FALSE, warning = FALSE}
library("tidyverse")
library("psych")
library("mice")
library("micemd")
library("miceadds")
library("mitml")
library("Amelia")
library("jomo")
library("parallel")
library("future")
library("furrr")
library("nlme")
library("MplusAutomation")
```

## Data

```{r}
data(Oxboys, package = "nlme")

Oxboys_addNA <- data.frame(Oxboys)
Oxboys_addNA <- Oxboys_addNA %>% 
  rename(id = Subject)

Oxboys_addNA$id <- as.integer(Oxboys_addNA$id)
Oxboys_addNA$Occasion <- as.integer(Oxboys_addNA$Occasion)

set.seed(52242)
Oxboys_addNA[sample(1:nrow(Oxboys_addNA), 25), "height"] <- NA

dataToImpute <- Oxboys_addNA
varsToImpute <- c("height")
Y <- varsToImpute
```

# Approaches to Multiple Imputation

- multiple imputation by joint modeling
    - assumes that the variables in follow a joint distribution, e.g.:
        - multivariate normal (i.e., multivariate normal imputation)
    - R packages:
        - `jomo`
        - `pan`
        - `Amelia`
        - `Mplus`
- multiple imputation by chained equations
    - aka:
        - fully conditional specification multiple imputation
        - sequential regression multiple imputation
    - R packages:
        - `mice`
            - predictive mean matching (`?mice::mice.impute.pmm`) can be useful to obtain bounded imputations for non-normally distributed variables

# Describe Missing Data

```{r}
describe(dataToImpute)
md.pattern(dataToImpute, rotate.names = TRUE)
```

# Pre-Imputation Setup

## Specify Number of Imputations

```{r}
numImputations <- 5 # generally use 100 imputations; this example uses 5 for speed
```

## Detect Cores

```{r}
numCores <- parallel::detectCores() - 1
```

# Multilevel Multiple Imputation

## Three-Level and Cross-Classified Data

https://simongrund1.github.io/posts/multiple-imputation-for-three-level-and-cross-classified-data/

## Methods

### `mice`

#### Types

##### Continuous Outcomes

https://stefvanbuuren.name/fimd/sec-multioutcome.html#methods

```{r, eval = FALSE}
?mice::mice.impute.2l.norm
?mice::mice.impute.2l.pan
?mice::mice.impute.2l.lmer
?miceadds::mice.impute.2l.pmm
?miceadds::mice.impute.2l.contextual.pmm
?miceadds::mice.impute.2l.continuous
?micemd::mice.impute.2l.2stage.norm
?micemd::mice.impute.2l.2stage.pmm
?micemd::mice.impute.2l.glm.norm
?micemd::mice.impute.2l.jomo
```

##### Discrete Outcomes

https://stefvanbuuren.name/fimd/sec-catoutcome.html#methods-1

```{r, eval = FALSE}
?mice::mice.impute.2l.bin
?miceadds::mice.impute.2l.binary
?micemd::mice.impute.2l.2stage.bin
?micemd::mice.impute.2l.glm.bin
```

#### Specifying the Imputation Method

```{r}
meth <- make.method(dataToImpute)
meth[1:length(meth)] <- ""
meth[Y] <- "2l.pmm" # specify the imputation method here; this can differ by outcome variable
```

#### Specifying the Predictor Matrix

A predictor matrix is a matrix of values, where:

- columns with non-zero values are predictors of the variable specified in the given row
- the diagonal of the predictor matrix should be zero because a variable cannot predict itself

The values are:

- NOT a predictor of the outcome: `0`
- cluster variable: `-2`
- fixed effect of predictor: `1`
- fixed effect and random effect of predictor: `2`
- include cluster mean of predictor in addition to fixed effect of predictor: `3`
- include cluster mean of predictor in addition to fixed effect and random effect of predictor: `4`

```{r}
pred <- make.predictorMatrix(dataToImpute)
pred[1:nrow(pred), 1:ncol(pred)] <- 0
pred[Y, "id"] <- (-2) # cluster variable
pred[Y, "Occasion"] <- 1 # fixed effect predictor
pred[Y, "age"] <- 2 # random effect predictor
pred[Y, Y] <- 1 # fixed effect predictor

diag(pred) <- 0
pred
```

#### Syntax

```{r}
mi_mice <- mice(
  as.data.frame(dataToImpute),
  method = meth,
  predictorMatrix = pred,
  m = numImputations,
  maxit = 5, # generally use 100 maximum iterations; this example uses 5 for speed
  seed = 52242)
```

### `jomo`

```{r}
level1Vars <- c("height")
level2Vars <- c("v3","v4")
clusterVars <- c("id")
fullyObservedCovariates <- c("age","Occasion")

set.seed(52242)

mi_jomo <- jomo(
  Y = data.frame(dataToImpute[, level1Vars]),
  #Y2 = data.frame(dataToImpute[, level2Vars]),
  X = data.frame(dataToImpute[, fullyObservedCovariates]),
  clus = data.frame(dataToImpute[, clusterVars]),
  nimp = numImputations,
  meth = "random"
)
```

### `Amelia`

- `!` in the console output indicates that the current estimated complete data covariance matrix is not invertible
- `*` in the console output indicates that the likelihood has not monotonically increased in that step

```{r}
boundVars <- c("height")
boundCols <- which(names(dataToImpute) %in% boundVars)
boundLower <- 100
boundUpper <- 200

varBounds <- cbind(boundCols, boundLower, boundUpper)

set.seed(52242)

mi_amelia <- amelia(
  dataToImpute,
  m = numImputations,
  ts = "age",
  cs = "id",
  polytime = 1,
  intercs = TRUE,
  #ords = ordinalVars,
  #bounds = varBounds,
  parallel = "snow",
  #ncpus = numCores,
  empri = .01*nrow(dataToImpute)) # ridge prior for numerical stability
```

### `Mplus`

#### Save `Mplus` Data File

Save R object as `Mplus` data file:

```{r}
prepareMplusData(dataToImpute_mplus, file.path(path, "dataToImpute_mplus.dat"))
```

#### `Mplus` Syntax for Multilevel Imputation

`Mplus` syntax for multilevel imputation:

```Mplus
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
!!!!!  MPLUS SYNTAX LINES CANNOT EXCEED 90 CHARACTERS;
!!!!!  VARIABLE NAMES AND PARAMETER LABELS CANNOT EXCEED 8 CHARACTERS EACH;
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

TITLE: Model Title

DATA: FILE = "datafile.dat";

VARIABLE:
  NAMES = id age height Occasion;
  MISSING = .;
  USEVARIABLES ARE age height Occasion;
  !CATEGORICAL ARE INSERT_NAMES_OF_CATEGORICAL_VARIABLES_HERE;
  CLUSTER = id;

ANALYSIS:
  TYPE = twolevel basic;
  ESTIMATOR = bayes;
  bseed = 52242;
  PROCESSORS = 2;
    
DATA IMPUTATION:
  IMPUTE = age height Occasion; !put ' (c)' after categorical vars
  NDATASETS = 100;
  SAVE = imp*.dat
```

# Parallel Processing

https://www.gerkovink.com/miceVignettes/futuremice/Vignette_futuremice.html

```{r}
mi_parallel_mice <- futuremice(
  dataToImpute,
  method = meth,
  predictorMatrix = pred,
  m = numImputations,
  maxit = 5, # generally use 100 maximum iterations; this example uses 5 for speed
  parallelseed = 52242,
  n.core = numCores)
```

# Imputation Diagnostics

## Trace Plots

On convergence, the streams of the trace plots should intermingle and be free of any trend (at the later iterations).
Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence.

- https://stefvanbuuren.name/fimd/sec-algoptions.html

```{r}
plot(mi_mice, c("height"))
```

## Density Plots

```{r}
densityplot(mi_mice)
densityplot(mi_mice, ~ height)
```

## Strip Plots

```{r}
stripplot(mi_mice)
stripplot(mi_mice, height ~ .imp)
```

# Post-Processing

## Modify/Create New Variables

### `mice`

```{r}
mi_mice_long <- complete(
  mi_mice,
  action = "long",
  include = TRUE)

mi_mice_long$newVar <- mi_mice_long$age * mi_mice_long$height
```

### `jomo`

```{r}
mi_jomo <- mi_jomo %>% 
  rename(height = dataToImpute...level1Vars.)

mi_jomo$newVar <- mi_jomo$age * mi_jomo$height
```

### `Amelia`

```{r}
for(i in 1:length(mi_amelia$imputations)){
  mi_amelia$imputations[[i]]$newVar <- mi_amelia$imputations[[i]]$age * mi_amelia$imputations[[i]]$height
}
```

## Convert to `mids` object

### `mice`

```{r}
mi_mice_mids <- as.mids(mi_mice_long)
```

### `jomo`

```{r}
mi_jomo_mids <- miceadds::jomo2mids(mi_jomo)
```

### `Amelia`

```{r}
mi_amelia_mids <- miceadds::datlist2mids(mi_amelia$imputations)
```

## Export for `Mplus`

```{r, eval = FALSE}
mids2mplus(mi_mice_mids, path = file.path("InsertFilePathHere", fsep = ""))
mids2mplus(mi_jomo_mids, path = file.path("InsertFilePathHere", fsep = ""))
mids2mplus(mi_amelia_mids, path = file.path("InsertFilePathHere", fsep = ""))
```

# Resources

https://stefvanbuuren.name/fimd

https://www.gerkovink.com/miceVignettes/Multi_level/Multi_level_data.html
